---
editor: 
  markdown: 
    wrap: sentence
---

# Traffic Characterisation

```{r, echo=FALSE}
#| label: setup
#| echo: false
#| message: false

source(here::here("R","03-traffic_characterization-graphs.R"))
```

To facilitate operational comparisons, it is crucial to have a good understanding of the level and composition of air traffic.
The preceding section provided an overview of the context and organisation of air navigation services in Brazil and Europe.
This chapter presents some air traffic characteristics for both regions to provide a framework for the observed operational performance in subsequent parts of the report.

## Network Level Air Traffic


@fig-annual-traffic-timeline shows the regional traffic development in Brazil and Europe.
In both regions the unprecedented decline in air traffic occurred in March 2020 in the aftermath of the pandemic-declaration by the World Health Organisation.
However, there is a difference in terms of the overall recovery.
The European recovery is characterised by two waves, while a single setback is observed in Brazil in second quarter of 2021.
The European pattern demonstrates the difficulty in the coordination of a joint policy of curbing the pandemic and managing travel related constraints.
With different states in Europe introducing public health and travel constraints at differnt times, intra-European traffic was affected by the piece-meal approach.
Brazil - and its policy on air transport - benefitted from single stance on policy implementation.

:::{callout-note}
TO BE CLEANED LATER

* reduce number of labels for BRA
* add associated labels to EUR

:::

```{r}
#| label: fig-annual-traffic-timeline
#| fig-cap: Regional daily air traffic
#| warning: false

plot_bra_annual_traffic_function() / plot_eur_annual_traffic_function()
```

::: callout-note
Brazilian air traffic continues to exceed 2019 levels, marking a phase of real growth, not just post-pandemic recovery. Compared to Europe, where traffic is approaching pre-pandemic levels, Brazil presents a different scenario. Since 2022, traffic has grown steadily, showing a structured recovery in demand. However, commercial aviation is still adjusting its route network in some regions to fully match 2019 levels.
The chart showing the daily evolution of air traffic (7-day moving average) highlights the busiest and quietest periods of recent years. Over the past three years, the day with the lowest average traffic typically falls on the Wednesday after Carnival period. Meanwhile, the annual peak usually happens in late December, close to Christmas, reinforcing a well-established seasonal pattern.
Overall, despite fluctuations throughout the year, the general trend is upward, confirming that both Brazil and Europe are on a steady path of growth in the air transport sector.
It is important to note that the moving average from Brazil data only includes commercial flights, excluding general and military aviation, and is based on UTC time.

:::


For Brazil, it is important to remember that @fig-annual-traffic-timeline shows the aggregated movements per airport at the whole network level.
The shown total does not necessarily reflect the total number of flights.
Another important observation related to the data is that Brazil's number of airports served with the TATIC tool (Tower ATC System) has increased.
Despite raising the processed total daily flight number, this difference is mostly transparent for this study as these additional airports handle only a small number of movements on a day-to-day basis.

The movements already surpassed the 2019 levels for the Brazilian region, confirming some economic recovery in the market.
According to the CGNA (Brazilian Network Manager) assessment, general aviation is the leading actor in this frame.
The share of "Light" aircraft in the fleet mix observed at Brazilian airports and the prevailing airline traffic levels still below the 2019 traffic in the airlines' preferred airports help to confirm this thesis.

In terms of total network level air traffic, the European region is still lagging behind its pre-pandemic levels.
Other analyses showed that low-cost carriers recovered more agile than the classical mainline carriers.
The low-cost sector, thus, shows a higher numbers of operations than pre-pandemic as their financial model allowed for a more agile reaction in terms of staffing/crewing/servicing flights.
The higher share can also be explained by a side-effect of the national support measures for some of the mainline carriers.
These measures included freeing slots at major hubs and the reduction of domestic / short-haul operations.
Accordingly, the European network is characterised by a change in the level of connectivity and frequency of services between the different airports.


```{r}
#| label: fig-annual-network
#| fig-cap: Evolution of annual network traffic
#| warning: false

this_years = c(2022,2023,2024)

# data loaded in supporting file - lift to here
# i.e. tfc_bra, tfc_eur

tfc_bra |> mutate(REG = "BRA") |> 
  bind_rows(tfc_eur) |> 
  
  # plot combined timelines
  plot_timeline_per_year(this_years) + 
  
  # beautify combined plot
  scale_y_continuous(limits = c(0,NA)) + 
  theme(legend.position = "inside", legend.position.inside = c(0.9, 0.6)) +
  facet_wrap(.~ REG, ncol = 1, scales = "free_y") 
```

::: callout-note
 The analysis of annual air traffic trends between 2022 and 2024 shows a clear consolidation of seasonal patterns in both Brazil and Europe. In Brazil, the curve remains more stable throughout the year, reflecting a demand pattern that is less affected by strong seasonal changes. In contrast, Europe shows a more distinct pattern, with a significant increase in traffic during the summer months and a drop in winter—typical of the region’s tourism-driven dynamics.
In Brazil, May saw a noticeable drop in traffic volume, mainly due to heavy rainfall in the south of the country starting in late April. The floods led to the suspension of operations at Salgado Filho International Airport (SBPA), in Porto Alegre, which remained closed until October. This significantly affected domestic commercial aviation in the region and had an impact on the national air traffic network.
While Brazil shows less variation in traffic across the year, European traffic clearly peaks between June and September. This highlights the need for targeted strategies to manage capacity and optimize airport infrastructure, especially during periods of high demand.
In both cases, we observe consolidated growth trends, although at different paces. Europe continues to face challenges such as network adjustments and restrictions in some airspaces, while Brazil keeps progressing steadily, even when dealing with regional disruptions.

:::


-   For Brazil, traffic in the first half of 2023 exceeded the pre-pandemic traffic level
-   On a network level, the continual recovery of the traffic is on-going in Europe.
-   The first half of 2023 saw traffic levels at about 90-95% of the pre-pandemic network traffic and started following a classical *seasonal* pattern. However, traffic recovery in Europe is not at the same level for each pre-pandemic service/connection. The recovery also resulted in a light modification of network connectivity. The Russian invasion of Ukraine resulted in the closure of a significant portion of the airspace (about 20%). However, the overall impact of the closure on air traffic - on a network level - was relatively small.

This high-level network perspective shows that traffic in both regions is comparable to pre-pandemic levels.
It will be interesting to observe the further evolution and growth of air traffic.

## Airport Level Air Traffic

The previous section showed the air traffic development on the network level.
As airports represent nodes in this overall network, changes to the overall traffic situation will ripple down to the airport level.
This demand on terminal and airport air navigation services forms a substantial input to understand how the operational performance measures in this report developed over time for the selected study airports.
This reports looks in particular at the performance levels observed at 10 key airports in each region (c.f. scope)


```{r}
 #| label: fig-bra-apt-tfc
 #| fig.cap: Brazilian airport level traffic
 
fig_bra_apt_tfc(c(2022,2023,2024))
```


```{r}
#| label: fig-eur-apt-level-tfc
#| fig-cap: Europeaan airport level traffic
#| warning: false

# TODO - move data loading to here

#annual_tfc_apt |> 
#  fig_eur_apt_tfc(.years = c(2022,2023,2024))

p_share <- annual_share_apts_network_eur |> 
  filter(YEAR %in% 2022:2024) |> 
  mutate(YEAR = as.character(YEAR)) |>
  p_share_of_network() + 
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

p_col_annual <- annual_tfc_apt |> mutate(YEAR = as.character(YEAR)) |> get_p1_eur(c(2022,2023,2024))

# patch both plots
p_share + p_col_annual + plot_layout(widths = c(1, 4))
```

```{r}
.apt = "SBGR"
.years = c("2023","2024")
airport_per_year(.apt, .years)
```

::: callout-note

xxxxxx The suggestion is to insert the fifth europe airport eith 2023 and 2024 levels.named this airport on the text.

The figure shows the monthly evolution of traffic at Guarulhos International Airport (SBGR) during 2023 and 2024, which remains the top airport in Brazil. A steady increase in operations can be seen throughout 2024 compared to 2023, with July standing out as the month with the highest volume during the period analyzed. This trend reflects the continued strengthening of commercial aviation at Brazil’s main hub, both for domestic and international flights.
Guarulhos’ performance becomes even more relevant when compared to the busiest airports in Europe. The fifth busiest airport in Europe in 2024 had traffic levels similar to those of Guarulhos, highlighting the different realities of each region. This comparison helps illustrate the structural complexity and differences between the two systems. While Brazil concentrates much of its traffic in a few key airports like SBGR, Europe spreads its operations across a larger network of major airports, often with more robust infrastructure (such as more runways). Therefore, direct comparisons between the “top 1” airports in each region may not accurately reflect their unique operational contexts.

:::


Analyzing the movement of the leading Brazilian airports, it is evident that they were not responsible for the return to 2019 levels, considering that only Campinas - SBKP and Santos Dumont - SBRJ slightly exceeded 2019 levels.
This phenomenon can be explained by the greater difficulty of airlines in resuming activity in contrast to general/business aviation.
This last share is an essential component of the Brazilian air movement but is a rare user of main airports.


```{r}
#| label: fig-apt-annual-change
#| fig-cap: Annual traffic at study airports in 2022 and variation 2022/2019
#| fig-pos: H

# Parâmetros: (ANO_REFERENCIA, ANO_COMPARACAO)
annual_change_bra(2024,2023) / annual_change_eur(2024,2023)
```

::: callout-note
The 2024 air traffic analysis highlights the importance of understanding the structural differences between the Brazilian and European networks. While Guarulhos (SBGR) remains the busiest airport in Brazil, its traffic volume is similar to that of the fifth busiest airport in Europe. This shows that direct comparisons between the top airports in each region may not fully reflect their different realities. Airport operations are influenced by many factors, such as the number of runways, network structure, geography, and density, which make the operational complexity unique in each case.
In Brazil, the largest traffic reduction in 2024 occurred at Porto Alegre airport (SBPA). The airport was directly affected by major flooding in May and remained closed until October. As a result, many flights were redirected to other airports in southern Brazil—airports that are not included in the scope of this report and, therefore, are not part of the comparison with Europe.
Another important change was the increase in traffic at Galeão International Airport (SBGL), which rose to sixth place among Brazil’s busiest airports. This shift mainly happened because part of the flights originally planned for Santos Dumont Airport (SBRJ) were moved to SBGL due to restrictions applied at SBRJ. Consequently, SBRJ saw a decrease in traffic, while SBGL experienced a gradual increase in operations.
Additionally, the share of total traffic handled by the main airports has changed. From 2022 to 2024, there is a slight downward trend, suggesting that traffic is becoming more spread out across a wider range of airports. This may be the result of regional recovery or strategic adjustments made by airlines.
Altogether, the data shows the Brazilian air network’s ability to adapt to climate events and operational changes, reinforcing the importance of continuous monitoring to ensure the system’s resilience and efficiency.

:::


@fig-apt-annual-change shows the annual change of the traffic served at the study airports in 2022 and the associated change of the traffic levels comparing 2022 and 2019.

With Campinas (SBKP) and Rio de Janeiro (SBRJ), there are two study airports in Brazil that serviced a higher level of traffic in 2022 than in 2019.
Both airports are key nodes for the domestic traffic in Brazil.
Salvador (SBSV) ranged at the pre-pandemic level.
The other Brazilian airports have seen - on average - a decrease of 10-20% of traffic.
This suggests that the observed network level increase in movements is distributed across the Brazilian network and not focussed on the airports covered in this study.

The European airport level traffic - on average - ranged at 20% below the pre-pandemic levels.
Munich (EDDM) and Rome (LIRF) observed higher reductions.
With an overall weaker recovery of the air traffic demand across Europe in 2022, a similar pattern emerged.
The increased network level traffic (ranging about 10% under the pre-pandemic level) is distributed across the European network and other aerodrome connections.

## Peak Day Traffic

While the annual traffic provides insights in the total air traffic volume and the associated demand, it does not provide insights on the upper bound of achievable daily movement numbers.
The latter depends on demand, operational procedures and/or associated constraints, and the use of the runway system infrastructure.
The peak day traffic is determined as the 99th percentile of the total number of daily movements (arrivals and departures).
The measure represents thus an upper bound for comparison purposes.


## Peak Day

:::{callout-note}

THERE IS A LOT OF CODE AND WE WORK WITH INTERIM AND "FIX" FILES.

It is not transparent how these files / values are generated.
On a first level, I could not fix it.
What needs to be done:

* [] produce consistent daily traffic files in PBWG format for BRA and EUR
* [] rewrite functions to read the PBWG format and aggregate as desired
* [] if desired bind the BRA and EUR data set to build the graphics

== CLEAN UP CODE!

:::

```{r}
pk_day_bra <- read_csv("./data/BRA-airport-tfc-peakday-fix.csv", show_col_types = FALSE)
our_years  <- c(2023,2024)

pk_day_bra |> rename(PEAK_DAY_PCT = PK_DAY) |> pk_day_plot(c(2022:2024)) + labs(subtitle = "magic at hand - check")
```


```{r}
#| label: fig-peak-day
#| fig-cap: Airport peak daily traffic (2024)
#| message: false

# where is the data
#peak_day_pct_plot(2024)
peak_day_comb |> 
  rename(PEAK_DAY_PCT = PK_DAY) |> 
  left_join(bra_apts_names) |> 
  filter(YEAR == 2023) |> 
  
plot_peak_day_tfc(2023)
```

::: callout-note
The figure shows that the peak day traffic between 2023 and 2024 shows stability in operational volumes across most of Brazil’s major airports. Even so, some notable variations are observed: Guarulhos (SBGR), Galeão (SBGL), and Confins (SBCF) recorded increases in their peak day movements, reflecting greater capacity to absorb traffic and some redistribution within the national network. In contrast, Santos Dumont Airport (SBRJ) experienced a significant drop in peak traffic, in line with operational restrictions that reduced its capacity and led to the transfer of some flights to Galeão.
In the case of Porto Alegre (SBPA), although it was the airport most affected by the floods that began in May 2024, there is no significant drop in peak day traffic. This is because the indicator reflects the highest daily traffic volume during the year, which likely occurred either before May or after partial operations resumed in October.
::: 

@fig-peak-day shows the peak day traffic in 2022 for the study airports with reference to the number of runways.
A varied picture can be seen for Europe.
For European with more than 2 runways it needs to be noted that the runway system does not support independent operations of all available runways.
Thus, the serviced peak traffic is also impacted by the runway system configuration.

The measure signals the use of the available runway system.

```{}
#| label: fig-timeline-peak-day
#| fig-cap: Evolution of peak-day traffic at study airports
#| message: false

timeline_peak_day(c(2019:2024))
```

::: callout-note
The figure shows historical data from 2019 to 2024, providing a broader view of the recovery path in both regions. In Brazil, Guarulhos (SBGR) remains the busiest airport every year. Other airports—such as Congonhas (SBSP), Brasília (SBBR), Campinas (SBKP), and Confins (SBCF)—also show steady growth, with some already surpassing pre-pandemic levels.
In Europe, xxxxxxxxx
These insights show the importance of analyzing peak day traffic as a complementary indicator to average daily movements, especially during periods of recovery or operational adjustments. The comparison between Brazilian and European contexts highlights the need for assessments tailored to the reality of each network, considering the structural, operational, and geographic factors that directly affect airport responsiveness and capacity.
:::


The year-to-year change of the peak day traffic between 2019 and 2022 is shown in @fig-timeline-peak-day.
For the European study airports, Frankfurt (EDDF), Munich (EDDM), Paris (LFPG), and Rome (LIRF) experienced a higher drop of the daily peak traffic in comparison to 2019.
Despite the not yet fully recovered demand situation at London Gatwick (EGKK) and Zurich (LSZH) showed a moderate reduction of the daily peak traffic in 2022.
This suggest that airports with limited airport runway capacity managed

## Fleet Mix

```{r}
#| label: fig-fleet-mix
#| fig-cap: Fleet mix observed at the study airports in 2024
#| fig-height: 5

# data is loaded with supporting script - move / clean and functionalise
# script returns fleetmix for fm_apts_bra and eur

fm_apts <- bind_rows(fm_apts_bra, fm_apts_eur) |> 
  left_join(bind_rows(bra_apts_names, eur_apts_names), by = join_by(ICAO))

key_year <- 2024

plot_fleet_mix(fm_apts, key_year)
```

@fig-fleet-mix depicts the observed share of different wake turbulence categories (WTC) across the study airports in `r key_year`.
In both regions, "medium" aircraft types are the predominant aircraft type.
The fleet mix - and in particular the separation requirements between the different aircraft types - is an important influencing factor for the capacity and observed (and achievable) throughput.
In general, a larger proportions of heavy aircraft or aircraft with longer runway occupancy times may result in lower throughput due to the required larger wake turbulence separation or time spent on the runway.
The locally defined capacity values may therefore differ based on the predominant fleet mix and operational characteristics, and ultimately result in different observed peak movement numbers or influence surface and terminal operations.

In Brazil, a significant number of "light" types operated in 2023.
For example Salvador (SBSV) serviced about 20% of "light" types.
The major hubs, i.e.
São Paulo Guarulhos (SBGR), Rio de Janeiro Galeão (SBGL), and Campinas (SBKP) observed a share of 15-20% of "heavy" aircraft.
These airports serve also as destinations for international long-haul flights.

With the exception of Zurich (LSZH), the share of "light" types is negligible at the European study airport in 2023.
London Heathrow (EGLL), Paris Charles de Gaule (LFPG), and Frankfurt (EDDF) observed the highest shares of "heavy" types.

Within the European region - and its multitude of national hubs - a significant number of international long-haul flights is operated at the chosen study airports.
In Brazil, the highlighted airports, Guarulhos (SBGR), Galeão (SBGL), and Campinas (SBKP), play a major role in terms of international connectivity.
It follows that medium and light types are used predominantly for inter-reginal connections.
Based on the selected study airports, the underlying decentralised structure of the European network becomes more visible.
Due to the geo-political composition, airports serving capitals or representing a main national hub are more frequent in Europe.
This is in contrast to Brazil, where the international and heavy air traffic appears more centralised at 2-3 pre-dominant hubs.

```{r}
#| label: fig-fleetmix-timeline
#| fig-cap: Fleet mix change over time observed at study airports
#| fig-height: 6

fleet_mix_timeline(c(2019:2024))
```


::: callout-note
 The 2024 data confirms the dominance of the "medium" aircraft category at the airports analyzed in both Brazil and Europe. Fleet mix plays a key role in airport capacity, directly impacting traffic flow and operational efficiency. Generally, a higher share of "heavy" aircraft can reduce runway throughput due to wake turbulence separation requirements and longer landing and takeoff times.
In Brazil, the main international hubs—Guarulhos (SBGR), Galeão (SBGL), and Campinas (SBKP)—showed a 15% to 20% share of "heavy" aircraft, reinforcing their role as the country’s main international gateways. Additionally, some Brazilian airports such as Brasília (SBBR) and Salvador (SBSV) reported a significant share of "light" aircraft, a category that is almost absent at the European airports analyzed—except for Zurich (LSZH). In Salvador, "light" aircraft account for around 20% of movements, and the share is also notable in Campinas (SBKP).
In Europe, major hubs like Frankfurt (EDDF), Heathrow (EGLL), and Charles de Gaulle (LFPG) continue to operate with a higher share of "heavy" aircraft, consistent with their status as global connection points. The structural difference between the two regions remains clear: Europe has a more decentralized international network, with multiple hubs across different countries, while Brazil concentrates most of its long-haul traffic in a few key airports.
The historical fleet mix chart shows that the distribution across aircraft categories has remained relatively stable over time. Even during the pandemic, although traffic volumes decreased, the mix of aircraft types operating at the analyzed airports did not change significantly. This indicates that all segments were proportionally impacted, without major structural shifts in the profile of operations.
Overall, the 2024 data confirms that Brazil continues to follow a more centralized model for international connectivity, while Europe operates with a broader, more distributed network. The presence of a large number of "light" aircraft in some Brazilian airports also highlights specific characteristics of regional aviation in the country, which influence traffic patterns and the use of airport infrastructure.
:::




On average, @fig-fleetmix-timeline shows that the fleetmix remained fairly stable over the years.
It is interesting to observe that the unprecedented decline in air transport during the pandemic phase did not substantially break this pattern.
This suggests that the contraction of the traffic volume hit all segments at a similar rate [^03-traffic_characterisation-1].

[^03-traffic_characterisation-1]: It must be noted that conceptually, the number of aircraft remained unchanged in the both regions.
    The higher utilisation of "heavy" aircraft for logistical support appeared to have offset the lower number international / long-haul passenger flights.
    For the first six month of 2023, this pattern is continued

## Summary

This chapter described the overall evolution of air traffic in Brazil and Europe on the network and study airport level.
Air traffic observed an unprecedented decline in both regions in response to COVID19.
However, the recovery path in both regions differed.
Overall, the Brazilian demand recovered substantially smoother than the European.
As national governments varied in their assessments and introduction of public health measures, including travel restrictions, intra-European air connectivity observed several setbacks.
On a network level, air traffic in Brazil exceeded the pre-pandemic level, whereas European traffic reaches about 5-10% below the pre-pandemic level.

In 2023, medium aircraft types were predominant in both regions across the studied airports.
Brazil hosted proportionally more light aircraft, notably in Salvador (SBSV).
At the same time, major hubs like São Paulo Guarulhos (SBGR) and Campinas (SBKP) handled substantial shares of heavy aircraft.
Conversely, European airports, except Zurich (LSZH), had minimal shares of light types, with London Heathrow (EGLL), Paris Charles de Gaule (LFPG), and Frankfurt (EDDF) witnessing higher shares of heavy types.
The European network demonstrated a decentralized structure with numerous international long-haul flights, while Brazil's international and heavy air traffic was centralized in 3 primary hubs.
Despite pandemic-induced declines in air transport, data suggests a stable fleet mix, indicating a consistent impact across traffic segments.

The overall air traffic demand situation is a key driver for the performance of air navigation services.
The observed differences may impact - inter alia - separation and synchronisation of air traffic, and influence the observed performance reported in the other chapters of this report.
