---
bibliography: references.bib
---

```{r}
#| echo: false
#| message: false
source(here::here("_chapter-setup.R"))
source(here::here("R","04-predictability_new-graphs.R"))
```

# Predictability

The preceding sections have demonstrated that both systems exhibit unique reactions to the broader developments in air transport.
Predictability plays a crucial role in these systems, impacting their functioning during both the strategic phase, where airline schedules are formulated, and the operational phase, where Air Navigation Service Providers (ANSPs) and stakeholders manage the delicate balance between demand and capacity.
Enhanced predictability stands to be advantageous for ANSPs, mainly when serving airspace users, as it contributes to highly efficient operations, even during periods of peak demand.
This chapter focuses on arrival and departure punctuality as crucial predictability indicators.


## Arrival Punctuality

The arrival punctuality shows the predictability of landing operations at airports, based on the scheduled in-block time (SIBT). It considers a 15-minute window for early or late arrivals and expresses the percentage of flights arriving at the gate within that margin.


```{r}
#| label: fig-arrival-punc
#| fig-cap: Evolution of arrival punctuality at study airports (2023 vs 2024)
#| fig-subcap: 
#|   - "Arrival punctuality 2023"
#|   - "Arrival punctuality 2024"
#| layout-nrow: 2


# debugging delay contributions
debug_arrs <- FALSE

# Chamadas da função
p_2023 <- create_punc_plot("ARR", 2023, debug_arrs)
p_2024 <- create_punc_plot("ARR", 2024, debug_arrs)

p_2023 / p_2024 + plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
```

::: callout note
Why @fig-arrival-punc label is (a) Arrival puctuality 2023??
::: 

@fig-arrival-punc shows the 2024 data and reaffirms key structural differences in punctuality behavior between Brazilian and European airports. Brazilian airports continue to report a high share of early arrivals—more than 15 minutes ahead of schedule—accounting for 20% to 30% of flights. In contrast, most European airports show early arrival shares below 15%. This pattern, observed consistently over recent years, reflects the use of built-in buffer times in Brazil’s scheduling practices. While these buffers help airlines improve on-time performance records, they can reduce predictability and complicate planning for air traffic management and airport operations.

Across both regions, the share of flights arriving within the -/+ 15 minute window remains the key measure of operational predictability. European patterns remained relatively stable in 2024, while in Brazil, punctuality varied significantly among airports.

```{r}
#| label: fig-arr-punc-comparison-apts
#| fig.cap: Arrival Punctuality at two comparison study airports 
#| 
comp_apts = c("SBGR","LPPT")
what_years = c("2023","2024")

punc_mov_month_bra <- month_tfc_bra |>
  left_join(tmp_month) |> mutate(PHASE = "ARR", REGION = "BRA")

punc_mov_month_eur <- punc_eur_month |> filter(PHASE == "ARR") |> 
  select(APT, YEAR, MONTH, REGION, PHASE, TOT_FLTS_MONTH = N_VALID, WITHIN_M15P15 ) |> 
  mutate(YEAR = as.character(YEAR))

monthly_punc_comparison <- bind_rows(
  punc_mov_month_bra, punc_mov_month_eur
)

punc_mov_month_plot(monthly_punc_comparison, what_years, comp_apts) +
  facet_wrap(.~ APT) +
  labs(subtitle = "per discussion develop barchart vs traffic change per year side by side")
```


To contextualize regional contrasts, @fig-arr-punc-comparison-apts presents a side-by-side comparison between Guarulhos (SBGR) and Lisbon (LPPT) between 2023 and 2024. Despite LPPT handling less than half the volume of flights compared to SBGR, it maintained punctuality levels close to 45% in October 2024. Overall, LPPT showed gradual improvement throughout the year. In contrast, SBGR remained below its 2023 performance until September, after which punctuality slightly improved. This late-year improvement indicates a potential shift but also highlights ongoing challenges.

This comparison underscores how operational structure, traffic complexity, and scheduling strategies directly influence punctuality outcomes. In Brazil, concentrated demand at a few major hubs—especially SBGR, the country's busiest airport—makes it harder to sustain performance within the target window.

```{r}
#' Function plotting punctuality side-by-side to traffic
#' 
#' Note: patchwork requires for a collection of fill/color legends their 
#' hard "identical"-ity; thus, we set color=, fill=, and key_glyph and labs!
#' 
plot_punc_vs_tfc <- function(.df_punc_share, .share_var = WITHIN_M15P15){
  
  p_punc_share <- .df_punc_share |> 
    ggplot() + 
    geom_col(aes(x = {{ .share_var }}, y = ICAO, fill = YEAR, color = YEAR)
              , position = position_dodge2(preserve = "single")
              , key_glyph = "point"
             ) + 
    scale_x_continuous(labels = scales::percent_format()) +
    labs(x = element_blank(), y = element_blank()
         ,color = element_blank(), fill = element_blank())
  
  p_punc_tfc <- .df_punc_share |> 
    ggplot(aes(x = N_VALID, y = ICAO) ) +
    geom_line(aes(group = ICAO)) +
    geom_point(aes(fill = YEAR, color = YEAR), key_glyph = "point") +
    scale_x_continuous(
      labels = scales::label_number(scale = 1e-3, suffix = "k")
      ) +
    labs(x = element_blank(), y = element_blank()
          ,color = element_blank(), fill = element_blank())
  
  p_punc_vs_tfc <- p_punc_share + p_punc_tfc + 
    plot_layout(widths = c(2,1), guides = "collect") &
    theme(legend.position = "top")
  
  return(p_punc_vs_tfc)
}

which_years <- 2023:2024

#----------- prep BRA ----------------------
arr_punc_bra <- punc_mov_month |> 
  mutate(DATE = paste0(YEAR, "-", sprintf("%02d", MONTH),"-01") |> 
                lubridate::as_datetime()
         ) |> 
  filter(PHASE == "ARR", lubridate::year(DATE) %in% which_years)

arr_punc_bra <- arr_punc_bra |> 
  filter(year(DATE) %in% which_years) |> 
  select(ICAO = APT, DATE, PHASE, N_VALID = TOT_FLTS_MONTH, WITHIN_M15P15) |>
  punc_aggregate("ARR","year")

#---------- prep EUR -----------------------
arr_punc_eur <- punc_eur_data |> 
  filter(year(DATE) %in% which_years) |> 
  punc_append_slots() |> select(ICAO, DATE, PHASE, N_VALID, WITHIN_M15P15) |>  
  punc_aggregate("ARR", "year") 
```

```{r}
#| label: fig-punc-vs-tfc
#| fig-cap: Evolution of arrival punctuality -/+15 min vs traffic
arr_punc_bra |> filter(ICAO %in% bra_apts) |> 
  mutate( YEAR          = year(DATE) |> as.character()
       #  ,WITHIN_M15P15 = WITHIN_M15P15 / N_VALID
         ) |> 
  plot_punc_vs_tfc()

arr_punc_eur |> filter(ICAO %in% eur_apts) |> 
  mutate( YEAR          = year(DATE) |> as.character()
         ,WITHIN_M15P15 = WITHIN_M15P15 / N_VALID
         ) |> 
  plot_punc_vs_tfc()
```

::: callout note
renamed figure 5.3 and 5.4
Figure 5.3: Evolution of BRA - arrival punctuality -/+15 min vs traffic
Figure 5.4: Evolution of EUR - arrival punctuality -/+15 min vs traffic
cHECK BRA PUNCTUALITY ON FIGURE 5.3 AND INSERT TEXT

::: 


@fig-punc-vs-tfc presents a side-by-side view of airport-level arrival punctuality and movement evolution between 2023 and 2024.

Overall, European airports demonstrated improvements in arrival punctuality. Among them, Madrid (LEMD) and Barcelona (LEBL) recorded the largest increase in traffic volumes, though this was accompanied by a slight decline in punctuality performance. In contrast, both Amsterdam (EHAM) and London Gatwick (EGKK) not only saw significant traffic growth but also achieved the most notable improvements in arrival punctuality among all airports in the study.

Munich (EDDM), however, registered the sharpest drop in punctuality, despite an increase in flight movements. This reinforces two critical operational dynamics:
First, the cumulative nature of delays, which highlights the limited resilience of the flight network—when a single flight is delayed, subsequent flights tend to be impacted due to tight scheduling and lack of flexibility.
Second, increasing demand often exacerbates existing local resource constraints, especially in areas like passenger processing and turnaround operations. As a result, reactionary delays accumulate and propagate throughout the system, further reducing predictability and punctuality.


The previous section highlighted the overall arrival punctuality observed at the study airports. 
Lower levels of punctuality can negatively impact predictability of operations and thus put a stronger strain on resources managing the arrival flow. 
Next to the arrival airspace capacity, stronger variations of the scheduled arrival times pose challenges for the surface management, as taxi operations, including stand allocation and availability, might result in changes of the taxi patterns, queuing within the taxiway and apron/stand system.
@fig-early-vs-late-arrivals shows the share of flights arriving more than 15 minutes before and after their scheduled arrival time.

```{r}
#| label: fig-early-vs-late-arrivals
#| fig-cap: Change of share of early and late arrivals (2023 vs 2024)

# note: limits allows to set left and right max ranges
# keep same limits for BRA and EUR to allow for comparison
# .limits defaults to c(-0.5, 0.5) ~ 50%

comb_punc <- bind_rows(punc_bra, punc_eur)

p2023<-create_earlylate_plot(comb_punc, "ARR",2023, .limits = c(-0.3, 0.45)) + 
  facet_wrap(. ~ REGION, scales = "free_y")
p2024<-create_earlylate_plot(comb_punc, "ARR",2024, .limits = c(-0.3, 0.45)) + 
  facet_wrap(. ~ REGION, scales = "free_y")


(p2023 / p2024) + plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
```

@fig-early-vs-late-arrivals compares the share of early and late arrivals at each study airport, considering arrivals more than 15 minutes ahead of or behind schedule (-15 min or +15 min) in 2023 and 2024.
From a broader perspective, air traffic in Brazil continues to show a tendency toward early arrivals, while in Europe, delayed arrivals are more prevalent.

Guarulhos (SBGR) remained the Brazilian airport with the highest share of early arrivals in 2024, followed by Campinas (SBKP), both with more than 30% of flights landing ahead of schedule. As key hubs in Brazil’s network, this anticipation may reflect a deliberate strategy by airlines to better manage connections and mitigate delays within their own schedules.
However, from a flow management perspective, this lack of precision poses operational challenges, as it complicates the allocation of resources and the sequencing of arrivals within controlled airspace and on the ground.

In Europe, Madrid (LEMD) recorded the highest share of early arrivals in 2024, reaching nearly 20%. Still, this figure remains below the levels seen in Brazil, reinforcing the structural and strategic differences in scheduling practices and performance expectations between the two regions.
Airport operators were identified as the major contributors to primary delays (ground handling, staff shortage), followed by ATFM delays.
However, the aforementioned reactionary effect was the main driver of knock-on delays [@CODA-2023] [^04-predictability-1].


[^04-predictability-1]: See CODA report at https://www.eurocontrol.int/publication/all-causes-delays-air-transport-europe-annual-2022.

## Departure Punctuality

The departure punctuality reflects the predictability of take-off operations at monitored airports. It is based on the comparison between the scheduled off-block time (SOBT) and the actual off-block time (AOBT), using a 15-minute tolerance window for early or late departures. The indicator expresses the percentage of flights that leave the gate within this time margin.

```{r}
#| label: fig-departure-punc
#| fig-cap: Evolution of departure punctuality at study airports (2023 vs 2024)
#| layout-nrow: 2
#| fig-subcap: 
#|   - Departure punctuality 2023
#|   - Departure punctuality 2024

# debugging delay contributions
debug_arrs <- FALSE

# Chamadas da função
p_2023 <- create_punc_plot("DEP", 2023, debug_arrs)
p_2024 <- create_punc_plot("DEP", 2024, debug_arrs)

p_2023 / p_2024 + plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
```

::: callout note
Why @fig-departure-punc label is (a) Departure puctuality 2023??
::: 

@fig-departure-punc presents the departure punctuality results for the study airports. 
The 2024 data shows that, overall, airports perform better in managing outbound traffic, although challenges remain.
Different from arrival punctuality, where Brazilian airports showed a high share of early arrivals and wide variability across the network—departure punctuality metrics appear comparatively stronger, especially in terms of flights departing within the punctuality (-/+ 15 min) time window.

One example is Santos Dumont Airport (SBRJ), where near to 50% of departures occurred 5 to 15 minutes ahead of schedule, reinforcing a consistent tendency toward early operations. On the other hand, Galeão Airport (SBGL) was one of the highest shares of delayed departures of flights leaving more than 15 minutes behind schedule.These two airports operate in close proximity and have undergone significant operational changes in recent years, which likely contribute to the contrasting performance. 

Moreover, seasonal weather conditions—such as summer thunderstorms, wind gusts, and windshear, especially in Brazil’s Southeast region, can disrupt operations, increase runway occupancy time, or require frequent runway changes, further affecting predictability.

In summary, while departure punctuality shows a slightly more favorable scenario compared to arrival operations, maintaining high performance levels still demands continuous improvements—especially at high-density airports. Operational planning, weather adaptation, and runway configuration remain key factors for enhancing outbound flight reliability.


::: callout note
Suggest to put the same graph as on arr, with SBGR and LPTT, and exclude fig-evolution-DEP-punc and fig-evolution-DEP-punc_by_month
::: 

```{r}
#| label: fig-evolution-DEP-punc
#| fig-cap: Evolution of departure punctuality window

# punc_evolution_lineplot("DEP",c(0.4, 1))
punc_evolution_lineplot(punc_bra ,"DEP",c(0.4, 1))
```




```{r}
#| label: fig-evolution-DEP-punc_by_month
#| fig-cap: Evolution of arrival punctuality window by month

.APT = c("SBBR","SBGR","SBGL")
.YEAR = 2024

#punc_per_month(.APT, .YEAR, "DEP", c(0.4, 1))

punc_monthly <- punc_bra_month

punc_per_month(punc_monthly, .APT, .YEAR, "DEP",c(0.4, 0.8))
```



The preceding section highlighted how the general traffic conditions in the previous years influenced the dependability of arrival schedules.
In this section, we assess the degree of departure punctuality measured as the difference between the scheduled (i.e. planned) departure versus the observed actual off-block time.
@fig-departure-punc shows the overall departure punctuality at Brazilian and European airports in 2019 compared to 2022.
Despite traffic levels in 2022 still ranging below their 2019 pre-pandemic levels, the departure punctuality in 2022 was - on average - lower than before COVID.

The difference in departure and arrival punctuality between 2022 and 2019 was significantly more pronounced for Europe indicating an increased strain on the turnaround processes.
There has been a significant increase in poor performance days, with departure punctuality falling below 50% and arrival punctuality dropping below 60%, occurring more frequently than in 2019.
On the Brazilian side, the Galeão airport (SBGL) observed the highest share of delayed departure flights.
It should be noted that the SBGL is the only airport with the Apron Control service directly provided by the airport.
Some inefficiency in the coordination between Tower and Apron or divergence at the indicator collection point for the location may be contributing to the observed performance.

Departure punctuality appeared slightly higher in Brazil in 2022 in comparison to 2019 and outperforms the punctuality levels observed in Europe. 
It is also notworthy, that in Brazil there is a higher share of flights blocking off between 15 to 5 minutes before their scheduled time.
Further research may help to clarify the factors driving this phenomenon.


```{r}
#| label: fig-early-vs-late-departures
#| fig-cap: Change of share of early and late departures (2023 vs 2024)

# note: limits allows to set left and right max ranges
# keep same limits for BRA and EUR to allow for comparison
# .limits defaults to c(-0.5, 0.5) ~ 50%

comb_punc <- bind_rows(punc_bra, punc_eur)

p2023<-create_earlylate_plot(comb_punc, "DEP",2023, .limits = c(-0.1, 0.6)) + 
  facet_wrap(. ~ REGION, scales = "free_y")
p2024<-create_earlylate_plot(comb_punc, "DEP",2024, .limits = c(-0.1, 0.6)) + 
  facet_wrap(. ~ REGION, scales = "free_y")

(p2023 / p2024) + plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
```

The change of the share of early and late departures in 2023 and 2024 is shown in @fig-early-vs-late-departures.

Both Brazilian and European airports demonstrate a positive operational behavior: early departures (before scheduled time) are relatively small in both regions, which is favorable for maintaining schedule stability and high punctuality.

In Brazil, the patterns observed in 2023 and 2024 are notably similar across the study airports, indicating consistent operational practices. Galeão (SBGL) was the airport with the highest share of late departures, reaching approximately 30% in both years. Despite this, most Brazilian airports maintained a balanced distribution, with a large proportion of flights departing within the -/+ 15-minute window.

In Europe, while the general behavior remains similar, some airports experienced a deterioration in departure punctuality. Munich (EDDM) saw an increase in late departures, with the share rising above 40% in 2024. This trend reflects localized challenges that continue to affect surface operations and overall network sequencing.

While surface movement operations appear stable at most airports, the growing share of late departures at specific locations signals the need for continuous monitoring and management interventions to assure the predictability of airport operations.


## Summary

Arrival and departure punctuality play an important role in terms of balancing demand and capacity. 

Arrival punctuality revealed distinct regional patterns. Brazilian airports continued to show a high share of early arrivals, largely due to the use of built-in buffer times in flight schedules. While this improves on-time performance metrics, it complicates air traffic management by reducing predictability. In contrast, European airports generally maintained lower shares of early arrivals and more stable punctuality performance across the study period. Operational comparisons, such as between Guarulhos (SBGR) and Lisbon (LPPT), highlighted the challenges faced by large, high-density hubs in sustaining punctuality under growing demand.

Departure punctuality showed slightly better performance compared to arrivals in both regions. Early departures remained relatively rare, supporting schedule stability. However, challenges persisted, particularly at airports like Galeão (SBGL) in Brazil and Munich (EDDM) in Europe, where late departures increased. These patterns emphasize how local operational constraints, weather disruptions, and surface management practices directly influence performance.

In both regions, maintaining high predictability levels remains critical to support efficient surface operations, arrival sequencing, and passenger experience. Continuous adaptation, proactive operational planning, and effective resource management are essential to sustain and improve predictability, especially as traffic demand continues to grow.

