---
bibliography: references.bib
---

```{r}
#| echo: false
#| message: false
source(here::here("_chapter-setup.R"))
source(here::here("R","04-predictability_new-graphs.R"))
```

# Predictability

The preceding sections have demonstrated that both systems exhibit unique reactions to the broader developments in air transport.
Predictability plays a crucial role in these systems, impacting their functioning during both the strategic phase, where airline schedules are formulated, and the operational phase, where Air Navigation Service Providers (ANSPs) and stakeholders manage the delicate balance between demand and capacity.
Enhanced predictability stands to be advantageous for ANSPs, mainly when serving airspace users, as it contributes to highly efficient operations, even during periods of peak demand.
This chapter focuses on arrival and departure punctuality as crucial predictability indicators.


## Arrival Punctuality

The arrival punctuality shows the predictability of landing operations at airports, based on the scheduled in-block time (SIBT). It considers a 15-minute window for early or late arrivals and expresses the percentage of flights arriving at the gate within that margin.


```{r}
#| label: fig-arrival-punc
#| fig-cap: Evolution of arrival punctuality at study airports (2023 vs 2024)
#| fig-subcap: 
#|   - "Arrival punctuality 2023"
#|   - "Arrival punctuality 2024"
#| layout-nrow: 2


# debugging delay contributions
debug_arrs <- FALSE

# Chamadas da função
p_2023 <- create_punc_plot("ARR", 2023, debug_arrs)
p_2024 <- create_punc_plot("ARR", 2024, debug_arrs)

p_2023 / p_2024 + plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
```

::: callout note
Why @fig-arrival-punc label is (a) Arrival puctuality 2023??
::: 

@fig-arrival-punc shows the 2024 data and reaffirms key structural differences in punctuality behavior between Brazilian and European airports. Brazilian airports continue to report a high share of early arrivals—more than 15 minutes ahead of schedule—accounting for 20% to 30% of flights. In contrast, most European airports show early arrival shares below 15%. This pattern, observed consistently over recent years, reflects the use of built-in buffer times in Brazil’s scheduling practices. While these buffers help airlines improve on-time performance records, they can reduce predictability and complicate planning for air traffic management and airport operations.

Across both regions, the share of flights arriving within the -/+ 15 minute window remains the key measure of operational predictability. European patterns remained relatively stable in 2024, while in Brazil, punctuality varied significantly among airports.

```{r}
#| label: fig-arr-punc-comparison-apts
#| fig.cap: Arrival Punctuality at two comparison study airports 
#| 
comp_apts = c("SBGR","LPPT")
what_years = c("2023","2024")

punc_mov_month_bra <- month_tfc_bra |>
  left_join(tmp_month) |> mutate(PHASE = "ARR", REGION = "BRA")

punc_mov_month_eur <- punc_eur_month |> filter(PHASE == "ARR") |> 
  select(APT, YEAR, MONTH, REGION, PHASE, TOT_FLTS_MONTH = N_VALID, WITHIN_M15P15 ) |> 
  mutate(YEAR = as.character(YEAR))

monthly_punc_comparison <- bind_rows(
  punc_mov_month_bra, punc_mov_month_eur
)

punc_mov_month_plot(monthly_punc_comparison, what_years, comp_apts) +
  facet_wrap(.~ APT) +
  labs(subtitle = "per discussion develop barchart vs traffic change per year side by side")
```


To contextualize regional contrasts, @fig-arr-punc-comparison-apts presents a side-by-side comparison between Guarulhos (SBGR) and Lisbon (LPPT) between 2023 and 2024. Despite LPPT handling less than half the volume of flights compared to SBGR, it maintained punctuality levels close to 45% in October 2024. Overall, LPPT showed gradual improvement throughout the year. In contrast, SBGR remained below its 2023 performance until September, after which punctuality slightly improved. This late-year improvement indicates a potential shift but also highlights ongoing challenges.

This comparison underscores how operational structure, traffic complexity, and scheduling strategies directly influence punctuality outcomes. In Brazil, concentrated demand at a few major hubs—especially SBGR, the country's busiest airport—makes it harder to sustain performance within the target window.

```{r}
#' Function plotting punctuality side-by-side to traffic
#' 
#' Note: patchwork requires for a collection of fill/color legends their 
#' hard "identical"-ity; thus, we set color=, fill=, and key_glyph and labs!
#' 
plot_punc_vs_tfc <- function(.df_punc_share, .share_var = WITHIN_M15P15){
  
  p_punc_share <- .df_punc_share |> 
    ggplot() + 
    geom_col(aes(x = {{ .share_var }}, y = ICAO, fill = YEAR, color = YEAR)
              , position = position_dodge2(preserve = "single")
              , key_glyph = "point"
             ) + 
    scale_x_continuous(labels = scales::percent_format()) +
    labs(x = element_blank(), y = element_blank()
         ,color = element_blank(), fill = element_blank())
  
  p_punc_tfc <- .df_punc_share |> 
    ggplot(aes(x = N_VALID, y = ICAO) ) +
    geom_line(aes(group = ICAO)) +
    geom_point(aes(fill = YEAR, color = YEAR), key_glyph = "point") +
    scale_x_continuous(
      labels = scales::label_number(scale = 1e-3, suffix = "k")
      ) +
    labs(x = element_blank(), y = element_blank()
          ,color = element_blank(), fill = element_blank())
  
  p_punc_vs_tfc <- p_punc_share + p_punc_tfc + 
    plot_layout(widths = c(2,1), guides = "collect") &
    theme(legend.position = "top")
  
  return(p_punc_vs_tfc)
}

which_years <- 2023:2024

#----------- prep BRA ----------------------
arr_punc_bra <- punc_mov_month |> 
  mutate(DATE = paste0(YEAR, "-", sprintf("%02d", MONTH),"-01") |> 
                lubridate::as_datetime()
         ) |> 
  filter(PHASE == "ARR", lubridate::year(DATE) %in% which_years)

arr_punc_bra <- arr_punc_bra |> 
  filter(year(DATE) %in% which_years) |> 
  select(ICAO = APT, DATE, PHASE, N_VALID = TOT_FLTS_MONTH, WITHIN_M15P15) |>
  punc_aggregate("ARR","year")

#---------- prep EUR -----------------------
arr_punc_eur <- punc_eur_data |> 
  filter(year(DATE) %in% which_years) |> 
  punc_append_slots() |> select(ICAO, DATE, PHASE, N_VALID, WITHIN_M15P15) |>  
  punc_aggregate("ARR", "year") 
```

```{r}
#| label: fig-punc-vs-tfc
#| fig-cap: Evolution of arrival punctuality -/+15 min vs traffic
arr_punc_bra |> filter(ICAO %in% bra_apts) |> 
  mutate( YEAR          = year(DATE) |> as.character()
       #  ,WITHIN_M15P15 = WITHIN_M15P15 / N_VALID
         ) |> 
  plot_punc_vs_tfc()

arr_punc_eur |> filter(ICAO %in% eur_apts) |> 
  mutate( YEAR          = year(DATE) |> as.character()
         ,WITHIN_M15P15 = WITHIN_M15P15 / N_VALID
         ) |> 
  plot_punc_vs_tfc()
```

::: callout note
renamed figure 5.3 and 5.4
Figure 5.3: Evolution of BRA - arrival punctuality -/+15 min vs traffic
Figure 5.4: Evolution of EUR - arrival punctuality -/+15 min vs traffic
cHECK BRA PUNCTUALITY ON FIGURE 5.3
::: 


European airports saw their share of punctual flights in 2022 decrease broadly compared to 2019, even with a more proportional distribution than the Brazilian system.
For European operators, there were two primary factors contributing the the lower performance in 2022. 
The most significant is the cumulative effect, showing that the network of flights has little ability to absorb the delay of one specific delayed flight.
The knock-on effect was amplified by the local resource constraints in terms of passenger and turn-around facilitation. 
The incrasing traffic demand posed challenges at many airports in Europe. Delayed arrivals accumulated further reactionary delay and ultimately passed the delay systematically on to next flights.
Further constraints were linked to air space and flow restrictions resulting from the geo-political conflict surrounding the Russian invasion in Ukraine.
On average, arrival delays of 15 minutes or more compared to the schedule ranged between 25-35% across the Europen study airports in 2022.


The previous section highlighted the overall arrival punctuality observed at the study airports. 
Lower levels of punctuality can negatively impact predictability of operations and thus put a stronger strain on resources managing the arrival flow. 
Next to the arrival airspace capacity, stronger variations of the scheduled arrival times pose challenges for the surface management, as taxi operations, including stand allocation and availability, might result in changes of the taxi patterns, queuing within the taxiway and apron/stand system.
@fig-early-vs-late-arrivals shows the share of flights arriving more than 15 minutes before and after their scheduled arrival time.

```{r}
#| label: fig-early-vs-late-arrivals
#| fig-cap: Change of share of early and late arrivals (2023 vs 2024)

# note: limits allows to set left and right max ranges
# keep same limits for BRA and EUR to allow for comparison
# .limits defaults to c(-0.5, 0.5) ~ 50%

comb_punc <- bind_rows(punc_bra, punc_eur)

p2023<-create_earlylate_plot(comb_punc, "ARR",2023, .limits = c(-0.3, 0.45)) + 
  facet_wrap(. ~ REGION, scales = "free_y")
p2024<-create_earlylate_plot(comb_punc, "ARR",2024, .limits = c(-0.3, 0.45)) + 
  facet_wrap(. ~ REGION, scales = "free_y")


(p2023 / p2024) + plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
```

@fig-early-vs-late-arrivals compares the share of early and late arrivals at each study airport in 2019 and 2022.
From a high-level perspective, air traffic tends to arrive well ahead of schedule in Brazil, while Europe observes a higher share of delayed arrivals.
Guarulhos (SBGR) remained the Brazilian airport with the highest share of early flights in 2022 (i.e. 33%), followed by Campinas (SBKP) with 30%.
Both airports are essential hubs in the country, and anticipation can be a consequence sought by air operators for better accommodation of the flight network.
However, for flow control, this lack of precision is equally problematic, affecting the optimal allocation of resources for the provision of air traffic control and flow service.
In turn, Madrid (LEMD) was the European element with the most significant share of early arrivals (i.e. 22%) in 2022. 
Pre-pandemic such a share was observed at London Heathrow in 2019.
These shares still range about 11% lower than the highest shares in Brazil.
The distorted nature of the European network in 2022 becomes apparent when observing the share of delayed flights.
For example services at London Gatwick (EGKK) faced a share of 39% of delayed flights.
Airport operators were identified as the major contributors to primary delays (ground handling, staff shortage), followed by ATFM delays.
However, the aforementioned reactionary effect was the main driver of knock-on delays [@CODA-2023] [^04-predictability-1].


[^04-predictability-1]: See CODA report at https://www.eurocontrol.int/publication/all-causes-delays-air-transport-europe-annual-2022.

## Departure Punctuality

```{r}
#| label: fig-departure-punc
#| fig-cap: Evolution of departure punctuality at study airports (2023 vs 2024)
#| layout-nrow: 2
#| fig-subcap: 
#|   - Departure punctuality 2023
#|   - Departure punctuality 2024

# debugging delay contributions
debug_arrs <- FALSE

# Chamadas da função
p_2023 <- create_punc_plot("DEP", 2023, debug_arrs)
p_2024 <- create_punc_plot("DEP", 2024, debug_arrs)

p_2023 / p_2024 + plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
```

```{r}
#| label: fig-evolution-DEP-punc
#| fig-cap: Evolution of departure punctuality window

# punc_evolution_lineplot("DEP",c(0.4, 1))
punc_evolution_lineplot(punc_bra ,"DEP",c(0.4, 1))
```



```{r}
#| label: fig-evolution-DEP-punc_by_month
#| fig-cap: Evolution of arrival punctuality window by month

.APT = c("SBBR","SBGR","SBGL")
.YEAR = 2024

#punc_per_month(.APT, .YEAR, "DEP", c(0.4, 1))

punc_monthly <- punc_bra_month

punc_per_month(punc_monthly, .APT, .YEAR, "DEP",c(0.4, 0.8))
```



The preceding section highlighted how the general traffic conditions in the previous years influenced the dependability of arrival schedules.
In this section, we assess the degree of departure punctuality measured as the difference between the scheduled (i.e. planned) departure versus the observed actual off-block time.
@fig-departure-punc shows the overall departure punctuality at Brazilian and European airports in 2019 compared to 2022.
Despite traffic levels in 2022 still ranging below their 2019 pre-pandemic levels, the departure punctuality in 2022 was - on average - lower than before COVID.

The difference in departure and arrival punctuality between 2022 and 2019 was significantly more pronounced for Europe indicating an increased strain on the turnaround processes.
There has been a significant increase in poor performance days, with departure punctuality falling below 50% and arrival punctuality dropping below 60%, occurring more frequently than in 2019.
On the Brazilian side, the Galeão airport (SBGL) observed the highest share of delayed departure flights.
It should be noted that the SBGL is the only airport with the Apron Control service directly provided by the airport.
Some inefficiency in the coordination between Tower and Apron or divergence at the indicator collection point for the location may be contributing to the observed performance.

Departure punctuality appeared slightly higher in Brazil in 2022 in comparison to 2019 and outperforms the punctuality levels observed in Europe. 
It is also notworthy, that in Brazil there is a higher share of flights blocking off between 15 to 5 minutes before their scheduled time.
Further research may help to clarify the factors driving this phenomenon.


```{r}
#| label: fig-early-vs-late-departures
#| fig-cap: Change of share of early and late departures (2023 vs 2024)

# note: limits allows to set left and right max ranges
# keep same limits for BRA and EUR to allow for comparison
# .limits defaults to c(-0.5, 0.5) ~ 50%

comb_punc <- bind_rows(punc_bra, punc_eur)

p2023<-create_earlylate_plot(comb_punc, "DEP",2023, .limits = c(-0.1, 0.6)) + 
  facet_wrap(. ~ REGION, scales = "free_y")
p2024<-create_earlylate_plot(comb_punc, "DEP",2024, .limits = c(-0.1, 0.6)) + 
  facet_wrap(. ~ REGION, scales = "free_y")

(p2023 / p2024) + plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
```

The change of the share of early and late departures in 2019 and 2022 is shown in @fig-early-vs-late-departures.
This presentation highlights the earlier observations.
On average, European airports observed a higher share of delayed departures by a factor of 2-3 in comparison to Brazil.
While the picture varies, the distored nature of the traffic in 2022 is visible in the higher share of delay departures in 2022 in Europe in comparison to the pre-pandemic levels in 2019.
This put a strain on the schedule stability across Europe. 
Local issues (e.g. reduced facilication capacity at airports) affected air traffic services in terms of surface movement, but also rippled into the network affecting the sequencing of arrivals and departure traffic.

## Summary

Arrival and departure punctuality play an important role in terms of balancing demand and capacity. 
Strong distortions of the schedule will ultimately ripple down into reactionary delay and require a higher effort for both the arrival flow and surface movement control.
Turnaround distortions affect therefore the planning accuracy of air navigation services and may lead to unwanted side-effects (e.g. longer sequencing and holding in the terminal airspace, passenger inconvenience due to longer taxi-times/blocked gates).

Distinct patterns were observed in both regions that cannot only be explained by the level of traffic recovery. 
On average, a higher share of flights arrives well ahead of schedule in Brazil. This pattern is largely unchanged when comparing 2019 and 2022. 
The ripple effect of the low preparedness level of European airports to address the returning traffic in 2022 is well documented.
This yielded a significant low performance in terms of departure punctuality exceeding levels observed in Brazil by a factor of 2-3.
Distortions of the local schedule can have knock-on effects on the air navigation service provision, both in terms of surface movement and arrival sequencing.
More research is needed to investigate and understand the underlying drivers and to what extent regional connectivity influences the observed patterns. 