# Efficiency

```{r}
source(here::here("_chapter-setup.R"))
source(here::here("R","06-efficiency-graphs.R"))
```

<!-- this is a comment ... verbatim copy from previous report -->

Operational efficiency is a critical component in assessing the management and execution of operations. It provides insights in the management of arrival and departure flows and the associated separation and synchronisation activities. 
Inefficiencies can have an impact on user operations in terms of delays or excessive fuel burn.
In light of the previous chapters it is therefore interesting to study how the available capacity was utilised to service demand during the different flight phases.

The measures reported in this comparison report are based on the observed travel time for surface operations (i.e. taxi-in and taxi-out) and during the arrival phase. 
These travel times are compared with an associated reference time for a group of flights showing similar operational characteristics. 
The determined difference (i.e. additional time) measures the level of inefficiency.
It must be noted that high performance operations will still yield a certain share of measured additional times. 
Operational efficiency is therefore aiming at minimising rather than eliminating these additional times as they cannot be zero.

## Additional Taxi-In Time

The additional taxi-in time measures the travel time of an arriving aircraft from its touchdown, i.e. the actual landing time, to its stand/gate position, i.e. actual in-block time).
This elapsed taxi-in time is compared to an anticipated reference time for aircraft arriving at the same runway and taxiing to the same (group of) stand/gate position(s).
Research showed that the taxi-times are not dependent on the type of aircraft.
The additional taxi-in time indicator provides a measure of the management of inbound surface traffic.

This report utilises another source for the movement times at Brazilian airports. 
Next to the actual taxi-times, the new data source provides also gate/stand information.
Accordingly, additional taxi-times can be now determined on a per-gate basis.
Previous studies did not support this higher level of granularity.
The reader needs therefore to bear in mind that the reported results and trends differ from previous reports which were based on an  airport-wide aggregation.
The latter may be influenced by the predominant runway system configuration and frequently used stand/parking positions.


### Annual Evolution of Additional Taxi-in Times



```{r}
#| label: fig-txit-annual
#| fig-cap: Additional taxi-in time [min/arr] (2023-2024)
#| fig-height: 4

fig_txit_annual(c(2023,2024))
```



The annual development of the average additional taxi-in times at the study airports is depicted by @fig-txit-annual.

The 2-minute threshold per arrival continues to serve as a practical reference point for evaluating taxi-in efficiency.

In Brazil, taxi-in performance in 2024 remained generally good, with most airports maintaining values close to or below 2 minutes. Notable exceptions were Congonhas (SBSP), which recorded the highest average additional taxi-in time, reaching approximately 3 minutes per arrival, followed closely by Guarulhos (SBGR), also nearing 3 minutes. Outside these two major hubs, all other studied Brazilian airports displayed taxi-in times well controlled around or below the 2-minute threshold. A particular highlight was Santos Dumont (SBRJ), which achieved a significant reduction in taxi-in times compared to the previous year.

In Europe, the patterns were slightly more varied. London Heathrow (EGLL) remained the airport with the highest additional taxi-in time, approaching 4 minutes per arrival in 2024, underlining persistent surface congestion challenges. Lisbon (LPPT) and Frankfurt (EDDF) also surpassed the 2-minute threshold, though only slightly. All other European airports maintained taxi-in times below 2 minutes. A remarkable highlight was Barcelona (LEBL), where taxi-in efficiency improved notably, with average additional taxi-in times dropping to below 1 minute per arrival.

Compared to 2023, taxi-in times remained relatively stable in both regions, although slight deteriorations were observed at some of the busiest hubs. These results emphasize the continued importance of optimizing surface management procedures to maintain overall predictability, particularly as traffic volumes increase at major airports.


### Monthly Variation of Additional Taxi-in Times

```{r}
#| label: fig-bra-txit-timeline
#| fig-cap: Evolution of average additional taxi-in time at Brazilian airports
#| fig-height: 4

txit_bra_per_month()
```

```{r}
#| label: fig-eur-txit-timeline
#| fig-cap: Evolution of average additional taxi-in time at European airports
#| fig-height: 4


txit_eur_per_month()
```


::: callout-note
is the graph @fig-eur-txit-timeline.  correct?
Looks like almost the same information

:::

The evolution of the taxi-in time at the study airports, shown in @fig-bra-txit-timeline and @fig-eur-txit-timeline, reinforces the findings presented in the previous analysis.
In Brazil, the figures clearly illustrate the increase in average taxi-in times at Porto Alegre (SBPA), Congonhas (SBSP), Salvador (SBSV), and Galeão (SBGL) throughout 2024. Congonhas and Guarulhos consistently maintained the highest levels within the Brazilian group, while Santos Dumont (SBRJ) stands out with a significant improvement, reducing its taxi-in times compared to previous years.

In Europe, the data reveal a relatively stable pattern over the two-year period. Most airports maintained consistent taxi-in times, with minor variations. Lisbon (LPPT) showed a slight upward trend, while Barcelona (LEBL) remained a highlight, sustaining very low taxi-in times, well below 1 minute per arrival.

These trends emphasize the importance of continuous monitoring of ground operations efficiency, especially as traffic demand grows.

## Taxi-Out Times

### Annual Evolution of Additional Taxi-out Times

```{r}
#| label: fig-txot-annual
#| fig-cap: Average additional taxi-out time [min/dep] (2023-2024)
#| fig-height: 4

plot_annual_txot(c(2023,2024))
```

On average, higher additional times for taxi-out are observed across all airports (c.f. @fig-txot-annual).

In Brazil, Santos Dumont (SBRJ) showed a notable improvement, with a reduction of more than 2 minutes compared to 2023. For the remaining airports, only minor variations were observed, including slight increases at Curitiba (SBCT) and Galeão (SBGL). Congonhas (SBSP) continues to record the highest additional taxi-out times among Brazilian airports, reaching approximately 7 minutes.

In Europe, a significant reduction in taxi-out times was observed across most airports. Zurich (LSZH) particularly stands out, achieving a reduction of about 2 minutes. Lisbon (LPPT) registered the highest taxi-out time among the European airports, nearing 7 minutes. Except for Lisbon, which saw a slight increase, all other European airports improved their taxi-out performance between 2023 and 2024.

### Monthly Variation of Additional Taxi-out Times

```{r}
#| label: fig-txot-monthly
#| fig-cap: Brazil - Monthly Evolution of taxi-out times
#| fig-height: 4

txot_bra_per_month()
```


```{r}
#| label: fig-txot-eur-monthly
#| fig-cap: Europe - Monthly Evolution of taxi-out times
#| fig-height: 4

txot_eur_per_month()
```



The trends reinforce the observations from the annual figures.

In Brazil, Santos Dumont (SBRJ) showed a significant improvement in 2024, with taxi-out times decreasing by more than two minutes. For the other airports, monthly variations were relatively small, although a gradual increase can be seen at Curitiba (SBCT) and Galeão (SBGL).
Congonhas (SBSP) continues to register the highest additional taxi-out times among Brazilian airports, reaching nearly 7 minutes in 2024.

In Europe, the overall trend is one of improvement, with most airports reducing their taxi-out times between 2023 and 2024. Zurich (LSZH) achieved a notable reduction of about two minutes. Lisbon (LPPT) stands out with the highest taxi-out time, close to 7 minutes. Except for Lisbon, which saw a slight increase, all other European study airports experienced reductions in taxi-out times, confirming a broad regional effort to improve surface efficiency.


## Mapping Additional Taxi-in and Taxi-out Times


```{r}
#| label: fig-txot-txit-mapping
#| fig-cap: Mapping of additional taxi-in and taxi-out times
  
plot_mapping_year1_year2(2023,2024)
```


::: callout-note

Is figure from 2019 / 2023 or 2023 / 2024 and only lables are not correct?
Based on this, the text was not updated

:::


This analysis builds on the previous sections. 
@fig-txot-txit-mapping compares the relationship between the taxi-in and taxi-out performance observed.

It also shows that on average taxi-out operations accrued more additional time than taxi-in operations (data points range below the dotted unit line, and as shown in the previous sections).

For most of the airports, the overall performance shows a reduction in additional taxi-out times (i.e. characterised by a leftshift along the x-axis).

A significant improvement in taxi-in performance can be obsered in Brazil across all study airports (i.e. decreasing trend along y-axis). 
This is contrasted by the behaviour in Europe. 
The majority of European airports observed no significant change in their taxi-in performance (i.e. no vertical trend).
The noteworthy exemption is London Heathrow (EGLL). 
EGLL faced a significant increase in average additional taxi-in time in 2023 in comparison to the pre-pandemic performance level observed in 2019.
The lower performance in terms of taxi-in is observed in @fig-eur-txit-timeline which shows a strong increase in the second half of 2023.
@fig-txot-txit-mapping also shows that the overall taxi-performance in Europe tends to show lower levels of variation between pre-pandemic and post-pandemic.
Exemptions are London Heathrow (EGLL) and Rome Fiumincino (LIRF) that saw a discernible rise in taxi-in inefficiencies and exceeding the levels observed at Brazilian study airports.


## Additional Time in Terminal Airspace

The additional time in terminal airspace is calculated as the difference of the actual flying time from entering the sequencing area (i.e. 100NM radius around the airport) to the actual landing time.
Previous research and guidance suggests that reference time can be build for flights sharing similar operational characteristics (entry sector, aircraft class, and landing runway).


```{r}
#| label: fig-asma
#| fig-cap: Additional time in terminal airspace
#| fig-height: 4

bra_eur_asma_plot(c(2023:2024))
```

::: callout-note
European data is not with 2024
The text not updated

:::
@fig-asma compares the annual average of additional times in terminal airspace across the study airports.
On average, the arrival flows at European airports are less constraint than in Brazil.

At London Heathrow (EGLL) a change in the operational concept helped to reduce the excessive additional ASMA times observed pre-COVID.
It must be noted that the minimal results for both London airports (Heathrow and Gatwick) in 2021 are a data artefact. 
The reference times for this study are build on the basis of the performance observed in 2019.
The lower traffic levels resulted in more efficient and shorter terminal sequencing operations. Compared to the 2019 reference times yields therefore minimal additional times.
Although there are improvements across the operations at European airports in comparison to 2019, it appears that with higher demand, procedural aspects applied before the pandmic are being reintroduced.

During the pandemic, Brazilian aviation authorities relieved regulations for general aviation in the São Paulo Congonhas (SBSP) airport. 
As this relief was still in force during the recovery in 2021 and 2022, balancing demand was difficult and caused higher additional time as the traffic increased back. 
For example, in May 2022, the lack of slots, associated with the gradual increase in demand, impacted the operations of Congonhas airport (SBSP). 
From June 2022, the need for slot for General aviation returned in force, but in October operations were again impacted by the amount of storms above expected for the season. 

For future reports, the data for for the assessment of the additional time in terminal airspace will be complemented for Brazil. 
This will allow to better investigate the changes observed across the years. [^BRA-asma]

[^BRA-asma]: The assessment of the additional time in terminal airspace  for Brazil is based on a data set starting in 2021. This may also result in reasonable small reference times and increase the measured values for Brazil. With a complemented dataset and a change of the base year for future reports, these data phenomena will be addressed.

```{r}
#| label: fig-asma-vs-traffic-volume
#| fig-cap: Comparison of additional time in terminal airspace

asma_vs_traffic_volume(c(2023,2024))
```

@fig-asma-vs-traffic-volume depicts the change in terms of the average additional time in terminal airspace comparing 2019 and 2022.
It must be noted that with respect to traffic, the Brazilian data set does not comprise all arriving traffic (c.f. traffic figures reported in earlier chapters).

The comparison shows the effect of the decline of air traffic on the performance in the European region. 
The contraction of demand resulted in lower observed additional times accrued by the arriving traffic.
For some airports in the Brazil region it can be observed how procedural aspects influence the additional time in terminal airspace.
For example, despite the variation of the traffic levels considered, the additional time remained fairly stable at SBGR comparing pre- and post-pandemic years.


## Summary

Operational efficiency provides an insight in terms of available benefit pools that can be exploited. 
In light of the ongoing climate change discussion, improvements in operational efficiency are directly linked with the aircraft flying time and can help to reduce unnecessary emissions.

Comparing the surface movement performance in Brazil and Europe shows similar trends.
On average, taxi-in operations are less constraint than taxi-out movements. 
The latter observed higher additional times. 
This may be subject to deconfliction of the traffic during the taxi-out phase, the establishment of the departure sequence, and holding at/close to the runway to maximise the runway utilisation.

The analysis of the additional time in terminal airspace on the basis of the new data set for Brazil confirmed previous trends.
On average, arrival sequencing in combination with the lower capacity result in higher additional times within the terminal airspace. 
It must be noted that Brazil undertook a major airspace redesign project in 2021 that influences the interface between the terminal airspace and the enroute network.
In the European region, the reduced level of air traffic resulted in lower pressure on the sequencing of arrivals.
However, the step increase between 2021 and 2022 for several airports suggests that constraints and more sequencing may come back with the increasing demand.

It will be interesting to study in future iterations of this comparison report to what extent arrival managment concepts deployed in the different regions contribute to the observed additional times.