---
title: "BRA-EUR - Towards Refining the Network Analysis"
format: pdf
---

```{r}
#| label: setup
#| message: false
#| warnin: false

source(here::here("_chapter-setup.R")) # all standard packages and defaults
```


# Overview

The idea is to develop a refined network analysis.    
For this we load the data for 2024 for Brazil (i.e. ANAC data set) and Europe (NM flight table).

```{r}
# -------------------- load the Brazilian network data ------------------------
cols <- c(
  "sg_empresa_icao",
  "nr_voo",
  "nm_mes_partida_real",
  "nr_mes_partida_real",
  "sg_icao_origem",
  "sg_icao_destino",
  "lt_combustivel",
  "sg_equipamento_icao",
  "nm_regiao_origem",
  "nm_pais_origem",
  "nm_regiao_destino",
  "nm_pais_destino",
  "nr_horas_voadas",
  NULL
)

bra_mvts <- read_csv("./data_proc/flights_ANAC_2024.csv"
                   #  , col_select = dplyr::any_of(cols)
                     , show_col_types = FALSE
                     ) |> 
  rename(
    ADEP = sg_icao_origem,
    ADES = sg_icao_destino,
    OPR  = sg_empresa_icao,            # aircraft operator (ICAO)
    FUEL_BURN = lt_combustivel,
    TYPE = sg_equipamento_icao,        # aircraft type (ICAO)
    REGIAO_DEP = nm_regiao_origem,
    DEP_CNTRY = nm_pais_origem,        # departing country
    REGIAO_ARR = nm_regiao_destino,
    DES_CNTRY = nm_pais_destino,       # destination country 
    FLT_DUR = nr_horas_voadas         # flight duration
)
```

```{r}
# ---------------- Europe --------------------------------------------------
# data is stored on local network drive
source(here::here("R","rqutils-zip.R"))
pth_to_data <- here::here(here::here() |> dirname(), "__DATA","NM-flight-table")
payload_zip <- "NM-flt-2024.zip"

check_zip_content(pth_to_data, payload_zip)  # check what we have in zip 

eur_mvts  <- read_zip(pth_to_data, payload_zip)
apts_meta <- arrow::read_parquet("./data/airport-icao-meta.parquet")

#| Utility function to prepare data -------------------------------------------
#| 
prep_eur_network_data <- function(.eur_mvts, .apts_meta = apts_meta){
  this_mvts <- .eur_mvts |> 
    dplyr::select(ADEP, ADES, FLTID = AIRCRAFT_ID, OPR = AIRCRAFT_OPERATOR
                  , REG = REGISTRATION, TYPE = AIRCRAFT_TYPE_ICAO_ID
                  , FLT_DUR = FLT_DUR_3
                  , DATE = LOBT
                  , SAM_ID = ID 
                  ) |> 
    dplyr::mutate(DATE = date(DATE))
  
  # augment airport details
  this_mvts <- this_mvts |> 
    dplyr::left_join(.apts_meta |> 
                       dplyr::select(ADEP = ICAO, DEP_CNTRY = COUNTRY_ISO)
                     , by = dplyr::join_by(ADEP)) |> 
    dplyr::left_join(.apts_meta |> 
                       dplyr::select(ADES = ICAO, DES_CNTRY = COUNTRY_ISO)
                     , by = dplyr::join_by(ADES))

  return(this_mvts)
}

eur_mvts <- eur_mvts |> dplyr::bind_rows() |> prep_eur_network_data()
  # all 2024 flights within EUR airspace ---------------------------
#  eur_mvts |> arrow::write_parquet(
#          "./data-prep-eur/EUR-network-connectivity-analytic-2024.parquet"
#)
  # write out study airport data -----------------------------------
# eur_mvts |> 
#   filter(ADEP %in% eur_apts | ADES %in% eur_apts) |> 
#   arrow::write_parquet(
#          "./data/EUR-network-connectivity-analytic-2024.parquet")
```




```{r}
calc_summary_routes_dep <- function(.mvts){
  this_summary <- .mvts |> 
    dplyr::group_by(ADEP) |> 
    dplyr::reframe(
       DEPS    = dplyr::n()
      ,NBR_DES = dplyr::n_distinct(ADES)) |> 
    dplyr::arrange(dplyr::desc(NBR_DES))
  
  return(this_summary)
}

bra_deps_routes <- bra_mvts |> calc_summary_routes_dep() |> mutate(REGION = "BRA")
eur_deps_routes <- eur_mvts |> bind_rows() |> 
  calc_summary_routes_dep() |> 
  left_join(oa |> select(ADEP = ICAO, CNTRY_ISO), by = join_by(ADEP)) |> 
  #------- patch unknown airports ------------------
  mutate(CNTRY_ISO = case_when(
     grepl(pattern = "^EB", x = ADEP) ~ "BE"
    ,grepl(pattern = "^E[DT]", x = ADEP) ~ "DE"
    ,grepl(pattern = "^EG", x = ADEP) ~ "GB"
    ,grepl(pattern = "^EH", x = ADEP) ~ "NL"
    ,grepl(pattern = "^EI", x = ADEP) ~ "IE"
    ,grepl(pattern = "^EN", x = ADEP) ~ "NO"
    ,grepl(pattern = "^EP", x = ADEP) ~ "PL"
    ,grepl(pattern = "^LE", x = ADEP) ~ "ES"
    ,grepl(pattern = "^LF", x = ADEP) ~ "FR"
    ,grepl(pattern = "^LG", x = ADEP) ~ "GR"
    ,grepl(pattern = "^LH", x = ADEP) ~ "HU"
    ,grepl(pattern = "^LI", x = ADEP) ~ "IT"
    ,grepl(pattern = "^LK", x = ADEP) ~ "CZ"
    ,grepl(pattern = "^LL", x = ADEP) ~ "IL"    # Israel
    ,grepl(pattern = "^LO", x = ADEP) ~ "AT"
    ,grepl(pattern = "^LR", x = ADEP) ~ "RO"
    #---------- other ADEPs ----------------------
    ,grepl(pattern = "^GM", x = ADEP) ~ "MA"    # Morocco
    ,.default = CNTRY_ISO
  )) |> 
  filter(CNTRY_ISO %in% (icao_iso_other |> filter(eurocontrol_pru %in% "Eurocontrol") |> pull(iso2c)) ) |> 
  mutate(REGION = "EUR")

# TODO - save as interim data and work from there
comb_deps_routes <- bind_rows(bra_deps_routes, eur_deps_routes) |> 
  group_by(REGION) |> 
  arrange(desc(DEPS), .by_group = TRUE) |> 
  mutate(RANK = 1:n(), SHARE = DEPS / sum(DEPS), CUM_SHARE = cumsum(SHARE)
         ,IN_STUDY = ADEP %in% c(bra_apts, eur_apts)) |> 
  ungroup()

#comb_deps_routes |> arrow::write_parquet("./data/BRA-EUR-dep-counts-2024.parquet")  
```

```{r}
if(exists("comb_deps_routes")){
  comb_deps_routes <- arrow::read_parquet("./data/BRA-EUR-dep-counts-2024.parquet")
}

#comb_deps_routes |> group_by(REGION) |> mutate(RANK = RANK / max(RANK)) |> 
comb_deps_routes |> filter(RANK <= 150) |> 
  ggplot() + 
  geom_step(aes(x = RANK, y = CUM_SHARE, group = REGION), direction = "hv")
```


```{r}
bind_rows(
   bra_deps_routes |> filter(ADEP %in% bra_apts) |> mutate(REGION = "BRA")
  ,eur_deps_routes |> filter(ADEP %in% eur_apts) |> mutate(REGION = "EUR") 
  ) |> 
  mutate(ADEP = fct_infreq(ADEP, NBR_DES) |> fct_rev()) |> 
  ggplot() +
  geom_col(aes(x = NBR_DES, y = ADEP, fill = REGION)) +
  scale_fill_manual(values = bra_eur_colours) +
  facet_wrap(.~ REGION, scales = "free_y") +
  labs(y = element_blank()) +
  theme(legend.position = "none")
```

```{r}
get_spread_of_destinations <- function(.deps){
  this_spread <- .deps |> 
    dplyr::group_by(ADEP, DES_CNTRY) |> 
    dplyr::reframe(
            DEPS = dplyr::n()
          , DEST = dplyr::n_distinct(ADES)) |> 
    dplyr::group_by(ADEP) |> 
    dplyr::mutate(
           SHARE_DEP      = DEPS / sum(DEPS)
         , TOP_DEST_SHARE = DEST / sum(DEST)
         ) |> 
    dplyr::arrange(desc(DEPS), .by_group = TRUE) |> 
    dplyr::ungroup()
  
  return(this_spread)
}

#------------------- BRA ----------------------------------------------
bra_dests <- bra_mvts |> 
  filter(ADEP %in% bra_apts) |> 
  get_spread_of_destinations()

#------------------ EUR --------------------------------------------------
eur_mvts_study <- arrow::read_parquet("./data/EUR-network-connectivity-analytic-2024.parquet")
ectrl_ms       <- readxl::read_xlsx("./data/ECTRL-ICAOEUR-state-memberships.xlsx") |>   filter(ECTRL == 1) |> select(STATE_ISO = `State-ISO`, STATE = State) |> 
  mutate(REGION = "EUR")

#------------ STATE and ISO CODES ----------------------------------------
global_iso2    <- ISOcodes::ISO_3166_1 |> 
  tibble() |> 
  select(STATE_ISO = Alpha_2, STATE = Name)

patch_global <- tibble::tribble(
  ~ STATE_ISO, ~STATE,
  "AN", "Netherlands Antilles",
  "XK", "Kosovo"      # need to think how to add to EUR region
)
global_iso2 <- global_iso2 |> bind_rows(patch_global)

# world wide codes - we may need this for generalising BRA data
# constructed wiht ChatGPT - not validated !!!!
icao_iso2 <- read_csv("./data/country-icao-iso-territory.csv")

oa_url <- "https://raw.githubusercontent.com/davidmegginson/ourairports-data/refs/heads/main/airports.csv"
oa <- readr::read_csv(oa_url, show_col_types = F) 
oa <- oa |> 
  dplyr::filter(!is.na(icao_code)) |> 
  dplyr::select(ICAO = icao_code, NAME = name, CNTRY_ISO = iso_country)

# patch oa
patch_oa <- tibble::tribble(
  ~ ICAO, ~ NAME, ~ CNTRY_ISO
  ,"SBCO", "tbd" ,"BR"
  ,"SBJD", "tbd", "BR"    
  ,"SBJI", "tbd", "BR"
  ,"SBMI", "tbd", "BR"
 ,"SBMT" , "tbd", "BR"
 ,"SBPO", "tbd", "BR"
 ,"SBRD" , "tbd", "BR"
 ,"SBSG" , "tbd", "BR"
 ,"SBSI", "tbd", "BR"
 ,"SBSO", "tbd", "BR"
,"SBUY", "tbd", "BR"
,"SBVC" , "tbd", "BR"
,'GQNN', NA , "MR"    # Maritania
,'SDAG', NA ,'BR'
,'SDIY', NA ,'BR'
,'SDJV', NA ,'BR'
,'SDLO', NA ,'BR'
,'SDVZ', NA ,'BR'
,'SIFC', NA ,'BR'
,'SIRI', NA ,'BR'
,'SIUA', NA ,'BR'
,'SIXE', NA ,'BR'
,'SJZA', NA ,'BR'
,'SN6L', NA ,'BR'
,'SNAB', NA ,'BR'
,'SNBA', NA ,'BR'
,'SNCL', NA ,'BR'
,"SNEB", NA, "BR"
,"SNIG" , NA, "BR"
,"SNJM"  , NA, "BR"
,"SNLN" , NA, "BR"
,"SNRJ"  , NA, "BR"
,"SNSM"  , NA, "BR"
,"SNSS"  , NA, "BR"
,"SNTS"  , NA, "BR"
,"SNWS"  , NA, "BR"
,"SNZR"  , NA, "BR"
,"SSBN"  , NA, "BR"
,"SSGG"  , NA, "BR"
,"SSQO"  , NA, "BR"
,"SSRS"  , NA, "BR"
,"SSVL"  , NA, "BR"
,"SWBE"  , NA, "BR"
,"SWKQ"  , NA, "BR"
,"SWYN"  , NA, "BR"
,"SYEC"  , NA, "GN"
)
oa <- oa |> dplyr::bind_rows(patch_oa)

# icao_iso_other <- countrycode::codelist |> 
#   select(iso2c, country.name.en, continent, eurocontrol_statfor, eurocontrol_pru, icao.region)
# icao_iso_other |> readr::write_csv("./data/country-icao-iso-etc.csv")
icao_iso_other <- readr::read_csv("./data/country-icao-iso-etc.csv", show_col_types = F)

# to-do patch
#   iso2c country.name.en continent eurocontrol_statfor eurocontrol_pru icao.region
#   <chr> <chr>           <chr>     <chr>               <chr>           <chr>      
# 1 SN    Senegal         Africa    NA                  NA              AFI   


#------------------------------------------------------ END ISO ----------

eur_dests <- eur_mvts_study |> 
  filter(ADEP %in% eur_apts) |> 
  get_spread_of_destinations() |> 
  #---------- append ECTRL / EUR ---------
  left_join(global_iso2 |> 
              rename(DES_CNTRY = STATE_ISO), by = join_by(DES_CNTRY)
            ) |>   
  left_join(ectrl_ms |> select(-STATE) |> 
              rename(DES_CNTRY = STATE_ISO) , by = join_by(DES_CNTRY)
            )
```

```{r}
p <- bra_dests |> 
  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines")
  ) 
p
p + theme(legend.position = "none")
```
international traffic

```{r}
bra_dests |> filter(DES_CNTRY != "BRASIL") |> 
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev())
```


```{r}
# DES_CNTRY ~ PAIS_DESTINO

bra_dests |> filter(DES_CNTRY != "BRASIL") |> 
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |> 
  
  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(fill = element_blank()) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines"),
    legend.position = "inside", legend.position.inside = c(0.75,0.25)
  ) 
```

```{r}
eur_dests |>
  filter(is.na(REGION)) |>
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |>

  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(fill = element_blank()) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines"),
    legend.position = "inside", legend.position.inside = c(0.75,0.25)
  )
```

Appendix 1 to the Directives to Regional Air Navigation Meetings and Rules of Procedure for their Conduct (Doc 8144-AN/874), comprises the following regions:

* Africa-Indian Ocean Region (AFI);
* Asia Region (ASIA);
* Caribbean Region (CAR);
* European Region (EUR);
* Middle East Region (MID);
* North American Region (NAM);
* North Atlantic Region (NAT);
* Pacific Region (PAC);
* South American Region (SAM).

```{r}
eur_world_shares <- eur_dests |>
  filter(is.na(REGION)) |>
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |> 
  left_join(icao_iso2 |> rename(DES_CNTRY = `ISO Alpha-2 Code`, ICAO_REG = `ICAO Region`) |> 
              select(DES_CNTRY, ICAO_REG)) |> 
  # aggregate by ICAO Region
  group_by(ICAO_REG) |> 
  reframe(DEPS = sum(DEPS), DEST = sum(DEST)) |> 
  mutate(REG_SHARES = DEPS / sum(DEPS))

eur_world_shares
```
```{r}
eur_world_shares |> 
  ggplot() +
  geom_col(aes(x = REG_SHARES, y = ICAO_REG, fill = ICAO_REG))
```

EUR ~  non-ECTRL Europe (probably need to take this out)



# Workflow Demo - Getting from Code to Functions

```{r}
calc_adep_counts <- function(.ds){
  my_count <- .ds |> 
  group_by(ADEP, DEP_CNTRY) |>  # Group by ADEP and PAIS_ORIGEM
  summarise(count = n(), .groups = "drop") |>  
  arrange(desc(count))
  
  return(my_count)
}



bra_mvts |> calc_adep_counts() 

plot_barchart <- function(.ds){
  
  p <- .ds |> 
  ggplot() +
  geom_col(aes(x = ADEP, y = count)) +
    facet_wrap(.~ REGION)
  
  return(p)
}

bra_demo <- bra_mvts |> calc_adep_counts() |> mutate(REGION = "BRA")
eur_demo <- bra_mvts |> calc_adep_counts() |> mutate(REGION = "EUR")

p1 <- bra_demo |> plot_barchart()
p2 <- eur_demo |> plot_barchart()

p1 / p2

bind_rows(bra_demo, eur_demo) |> plot_barchart()
```

The maximum departures seen at Brazilian airport account for `r bra_demo |> pull(count) |> max()`.





# Prepare analytic data for study

[1] "ADEP"      "ADES"      "FLTID"     "OPR"       "REG"       "TYPE"      "FLT_DUR"  
 [8] "DATE"      "SAM_ID"    "DEP_CNTRY" "DES_CNTRY"

```{r}
bra_mvts <- read_csv("./data_proc/flights_ANAC_2024.csv"
                   #  , col_select = dplyr::any_of(cols)
                     , show_col_types = FALSE
                     ) 
trim_ANAC <- function(.anac){
  trimmed_anac <- .anac |> 
    dplyr::select(
       ADEP    = sg_icao_origem
      ,ADES    = sg_icao_destino
      ,FLT_NBR = nr_voo
      ,OPR  = sg_empresa_icao           # aircraft operator (ICAO)
      ,REG  = ds_matricula 
      ,TYPE = sg_equipamento_icao       # aircraft type (ICAO)
      ,AOBT    = hr_partida_real
      ,AIBT    = hr_chegada_real
      ,FLT_DUR = nr_horas_voadas        # flight duration
      ,DATE    = dt_referencia
      ,id_basica                        # ANAC ID
      ,DEP_CNTRY_NAME = nm_pais_origem  # departing country
      ,DES_CNTRY_NAME = nm_pais_destino # destination country
    )
  return(trimmed_anac)
}

prep_bra_connectivity <- function(.trimmed_anac){
  bra_analytic <- .trimmed_anac |> 
    dplyr::mutate(
       FLTID = paste0(OPR, FLT_NBR)
    )
 return(bra_analytic)
}

append_iso_adep_ades <- function(
      .flt_adep_ades
    , .iso_look_up = oa |> select(ICAO, CNTRY_ISO)
    ){
  this_flt <- .flt_adep_ades |> 
    dplyr::left_join(
       .iso_look_up |> dplyr::rename(ADEP = ICAO, DEP_CNTRY = CNTRY_ISO)
      , by = dplyr::join_by(ADEP)
      ) |> 
    dplyr::left_join(
      .iso_look_up |> dplyr::rename(ADES = ICAO, DES_CNTRY = CNTRY_ISO)
      , by = dplyr::join_by(ADES))
  
  return(this_flt)
}

append_ades_regions <- function(
     .flt_ades_iso2
    ,.region_lookup = icao_iso_other |> 
      filter(!is.na(iso2c)) |> 
      select(  DES_CNTRY = iso2c
             , DES_ICAO  = icao.region   
             , DES_STATF = eurocontrol_statfor
             , DES_PRU = eurocontrol_pru)
){
  this_flt <- .flt_ades_iso2 |> 
    dplyr::left_join(.region_lookup, by = dplyr::join_by(DES_CNTRY))
  
  return(this_flt)
}

append_adep_regions <- function(
     .flt_adep_iso2
    ,.region_lookup = icao_iso_other |> 
      filter(!is.na(iso2c)) |> 
      select(  DEP_CNTRY = iso2c
             , DEP_ICAO  = icao.region   
             , DEP_STATF = eurocontrol_statfor
             , DEP_PRU = eurocontrol_pru)
){
  this_flt <- .flt_adep_iso2 |> 
    dplyr::left_join(.region_lookup, by = dplyr::join_by(DEP_CNTRY))
  
  return(this_flt)
}
```

```{r}
eur_mvts |> append_adep_regions() |> 
  append_ades_regions() |> filter(year(DATE) == 2024) |> 
  arrow::write_parquet(sink = here::here("data_proc","EUR-network-connectivity-2024.parquet"))

bra_mvts <- bra_mvts |> trim_ANAC() |> append_iso_adep_ades() |> 
  append_adep_regions() |> 
  append_ades_regions()

bra_mvts |> arrow::write_parquet(sink = here::here("data_proc","BRA-network-connectivity-2024.parquet"))
```


```{r}
bra_reg_tfc <- bra_mvts |> 
  mutate(DOMESTIC = DES_CNTRY == "BR") |> 
  group_by(DES_ICAO, DOMESTIC) |> 
  reframe(DEPS = n())

bra_reg_tfc |> mutate(LABEL = ifelse(!DOMESTIC, DES_ICAO, "local"), SHARE = DEPS / sum(DEPS)) |> mutate(LABEL = forcats::fct_reorder(LABEL, SHARE)) |> drop_na() |> ggplot() + geom_col(aes(x = SHARE, y = LABEL, fill = LABEL)) + scale_x_continuous(labels = scales::percent_format()) + labs(x = element_blank(), y = element_blank())
```


```{r}
library(ComplexUpset)

dests_bra_eur <- bra_mvts |> 
  filter(DES_PRU == "Eurocontrol", DEP_CNTRY == "BR") |> 
  group_by(ADEP, ADES) |> 
  reframe(DEPS = n()) |> 
  arrange(desc(DEPS)) |> 
  mutate(SHARE = DEPS / sum(DEPS), DUMMY = 1
         ,CUM_SHARE = cumsum(SHARE)) 

tmp <- bra_mvts |> 
  filter(DEP_CNTRY == "BR", DES_PRU == "Eurocontrol") |> 
  group_by(ADEP, ADES, DES_PRU, DEP_CNTRY) |> 
  reframe(DEPS = n()) |> 
  arrange(desc(DEPS)) |> 
  mutate(SHARE = DEPS / sum(DEPS), CUM_SHARE = cumsum(SHARE)) |> 
  mutate(IS_STUDY = ADES %in% eur_apts) |> 
  pivot_wider(names_from = ADES, values_from = IS_STUDY)

tmp <- tmp |> filter(CUM_SHARE <= 0.8)
tiks <- names(tmp)[-c(1:4)]
tmp |> upset(tiks)
```


another cool and long airport lookup

```{r}
ip2_url <- "https://github.com/ip2location/ip2location-iata-icao/blob/master/iata-icao.csv"

# github differentiates between "display" and "raw" version
# we download raw versions!
ip2_url <- "https://raw.githubusercontent.com/ip2location/ip2location-iata-icao/refs/heads/master/iata-icao.csv"


icao_ip2 <- readr::read_csv(ip2_url)
```


```{r}
bra_mvts <- arrow::read_parquet("data_proc/BRA-network-connectivity-2024.parquet")
eur_mvts <- arrow::read_parquet("data_proc/EUR-network-connectivity-2024.parquet")
```

international traffic

```{r}
bra_mvts_intnl <- bra_mvts |> 
  filter(DEP_CNTRY == "BR") |>  
  mutate(
     TFC_TYPE = ifelse(DEP_CNTRY == "BR" & DES_CNTRY == "BR", "regional", "international")
    ,IS_STUDY = ADEP %in% bra_apts) |> 
    group_by(DEP_CNTRY, IS_STUDY, DES_CNTRY, TFC_TYPE, DES_PRU) |> reframe(DEPS = n())  |> arrange(desc(DEPS))

# bra_mvts_intnl |> mutate(YEAR = 2024) |> write_csv("./data/BRA-network-conns-intnl.csv")

eur_mvts_intnl <- eur_mvts |> 
  filter(DEP_PRU == "Eurocontrol") |> 
  mutate(TFC_TYPE = ifelse(DEP_PRU == "Eurocontrol" & DES_PRU == "Eurocontrol", "regional","international"), IS_STUDY = ADEP %in% eur_apts) |> 
    group_by(DEP_CNTRY, IS_STUDY, DES_CNTRY, TFC_TYPE, DES_PRU) |> 
  reframe(DEPS = n()) |> 
  arrange(desc(DEPS))

#eur_mvts_intnl |> mutate(REGION = "EUR", YEAR = 2024) |> 
  write_csv("./data/EUR-network-conns-intnl.csv")
```


upset presentation

for upset plot we need to identify the group membership"S".

```{r}
library(ComplexUpset)

bra_neighbours <- tribble(
  ~ISO2, ~CNTRY_NAME
  , "AR", "Argentina"
  , "CL", "Chile"
  , "CO", "Columbia"
  , "UY", "Uruguay")

bra_upset <- bra_mvts |> 
  mutate(ID = row_number(), IS_STUDY = ifelse(ADEP %in% bra_apts, 1, 0)
         , CASE = paste0(DEP_CNTRY,"-",DES_CNTRY,"_",IS_STUDY) 
         ,NEIGHBOUR = case_when(
           DES_CNTRY %in% bra_neighbours$ISO2 ~ DES_CNTRY
          , .default = "Other")
         , DUMMY = 1 ) |> 
  
  filter(DES_PRU %in% c("Southern America"), DES_CNTRY != "BR") |> 
  select(ID,CASE, IS_STUDY, DEP_CNTRY, DES_CNTRY, DES_PRU, DUMMY, NEIGHBOUR) |>
  mutate(  STUDY_APT = ifelse(IS_STUDY, TRUE, FALSE)
         , NON_STUDY_APT = ifelse(!IS_STUDY, TRUE, FALSE)) |> 
  pivot_wider(id_cols = c(ID, CASE, DES_CNTRY, DES_PRU, STUDY_APT, NON_STUDY_APT)
              , names_from = NEIGHBOUR
              , values_from = DUMMY, values_fill = 0)

# make nice names
# look up (named) vector
bra_neighbours_vec <- deframe(bra_neighbours)
study_apt_label    <- c(STUDY_APT = "airport in study", NON_STUDY_APT = "non-study airport")
nice_name_vec     <- c(bra_neighbours_vec, study_apt_label)

bra_upset <- bra_upset |> 
  rename_with(.fn = ~ nice_name_vec[.x], .cols = any_of(names(nice_name_vec))) 

tiks <- names(bra_upset)[-c(1:4)]
#tiks <- c("STUDY_APT", "Argentina", "Chile", "Columbia", "Uruguay", "Other", "NON_STUDY_APT")


plot_bra_upset <- bra_upset |> upset(
    intersect = tiks
  , name = "connections"
  , base_annotations = list(
    # fine-tune presentation of Intersection size portion
    'Intersection size'=intersection_size(
            fill = bra_col,
            text_colors = c(
                on_background='black', on_bar='white'
            )
            ,text_mapping = aes(
              label = paste0(
                 !!get_size_mode('exclusive_intersection'),"\n"
                ,!!upset_text_percentage())
            )
            , text = list(size = 9 , size.unit = "pt")
        )
    # add ggplot layer for total numbers
        + annotate(
            geom='text', x=Inf, y=Inf,
            label=paste('total number \nof non-regional \nflights:', nrow(bra_upset)),
            vjust=1, hjust=1
        ) +
      labs(y = "departures")
  ),
  # fine-tune query part
  # queries=list(upset_query(set='airport in study', fill='blue'))
  # remove the sizes/query part ... not sure what it adds
  set_sizes = FALSE
)
plot_bra_upset
```

```{r}
plot_bra_upset |> ggsave(filename = "./figures/BRA-upset-national-connections.png", width = 8, height = 5, bg = "white")
```


BRA-EUR traffic

```{r}
study_br <- c("SBBR", "SBCF", "SBCT", "SBGL", "SBGR", "SBKP", "SBPA", "SBRJ", "SBSP", "SBSV")
study_eu <- c("EDDF", "EDDM", "EGKK", "EGLL", "EHAM", "LEBL", "LEMD", "LFPG", "LPPT", "LSZH")

# Named vectors for readable names
br_labels <- c(
  SBBR = "Brasília", SBCF = "Belo Horizonte", SBCT = "Curitiba",
  SBGL = "Galeão", SBGR = "Guarulhos", SBKP = "Campinas", SBPA = "Porto Alegre",
  SBRJ = "Santos Dumont", SBSP = "Congonhas", SBSV = "Salvador"
)

eu_labels <- c(
  EDDF = "Frankfurt", EDDM = "Munich", EGKK = "Gatwick", EGLL = "Heathrow",
  EHAM = "Amsterdam", LEBL = "Barcelona", LEMD = "Madrid", LFPG = "Paris CDG",
  LPPT = "Lisbon", LSZH = "Zurich"
)

# trim connections
eur_bra_conns <- eur_mvts |> 
  filter(DEP_PRU == "Eurocontrol", DES_CNTRY == "BR") |> 
  group_by(ADEP, ADES) |> reframe(DEPS = n()) |> arrange(desc(DEPS)) 

conns_threshold <- 90

major_nonstudy_eur <- eur_bra_conns |> 
  filter(!(ADEP %in%  study_eu)) |>   # only exclude study airports on the origin side
  group_by(ADEP)  |> 
  reframe(TOT_DEPS = sum(DEPS))  |> 
  filter(TOT_DEPS >= conns_threshold) %>%
  pull(ADEP)

major_nonstudy_bra <- eur_bra_conns |> 
  filter(!(ADES %in%  study_br)) |>   # only exclude study airports on the origin side
  group_by(ADES)  |> 
  reframe(TOT_DEPS = sum(DEPS))  |> 
  filter(TOT_DEPS >= conns_threshold) %>%
  pull(ADES)


eur_bra_grouped <- eur_bra_conns %>%
  mutate(
    ORIG_GROUP = case_when(
      ADEP %in% study_eu ~ eu_labels[ADEP],
      ADEP %in% major_nonstudy_eur ~ paste0(ADEP, " (Major Non-Study)"),
      TRUE ~ "Other Europe"
    ),
    DEST_GROUP = case_when(
      ADES %in% study_br ~ br_labels[ADES],
      ADES %in% major_nonstudy_bra ~ paste0(ADES, " (Major Non-Study)"),
      TRUE ~ "Other Brazil"
    )
  )

```

```{r}
library(ggalluvial)

ggplot(eur_bra_grouped, aes(axis1 = ORIG_GROUP, axis2 = DEST_GROUP, y = DEPS)) +
  geom_alluvium(aes(fill = ORIG_GROUP), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey80", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3.5) +
  scale_x_discrete(limits = c("Europe", "Brazil"), expand = c(.05, .05)) +
  labs(y = "Flights", title = "Main Europe–Brazil Air Connections") +
  theme_minimal()



ggplot(eur_bra_grouped,
       aes(axis1 = ORIG_GROUP, axis2 = DEST_GROUP, y = DEPS)) +
  geom_alluvium(aes(fill = ORIG_GROUP), width = 0.2, alpha = 0.8) +
  geom_stratum(width = 0.2, fill = "grey90", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)),
            size = 3, angle = 0, hjust = 0.5, vjust = 0.5) +
  scale_x_discrete(limits = c("Origin", "Destination"), expand = c(.05, .05)) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())

```

```{r}
ggplot(eur_bra_grouped, aes(axis1 = ORIG_GROUP, axis2 = DEST_GROUP, y = DEPS)) +
  geom_alluvium(aes(fill = ORIG_GROUP), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey90", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3.2) +
  scale_x_discrete(limits = c("Europe", "Brazil")) +
  labs(title = "Flight Flows from Europe to Brazil", y = "Flights") +
  theme_minimal()

```

```{r}
library(ggforce)

library(ggplot2)
library(ggforce)
library(dplyr)

# Make sure both ORIG_GROUP and DEST_GROUP are factors (important!)
flows_grouped <- eur_bra_grouped %>%
  mutate(
    ORIG_GROUP = ifelse(ORIG_GROUP %in% study_eu, , "Study", "Non-Study"),
    DEST_GROUP = ifelse(DEST_GROUP %in% study_br,  "Study", "Non-Study")
  )

ggplot(flows_grouped) +
  geom_parallel_sets(aes(
    ORIG_GROUP,
    DEST_GROUP,
    weight = DEPS,
    fill = ORIG_GROUP
  ), alpha = 0.8) +
  # geom_parallel_sets_axes(axis.width = 0.1, fill = "gray90") +
  # geom_parallel_sets_labels(angle = 0, size = 3) +
  # scale_x_discrete(labels = c("Origin", "Destination")) +
  theme_minimal() +
  labs(title = "Flight Flows from Europe to Brazil",
       subtitle = "Grouped by Origin and Destination Airport Categories",
       y = "Number of Departures")

```

```{r}
flows_grouped <-  eur_bra_grouped |> 
  mutate(
    ORIG_STUDY = ifelse(ADEP %in% eur_apts, "Study", "Non-Study"),
    DEST_STUDY = ifelse(ADES %in% bra_apts, "Study", "Non-Study")
)
flows_long <- flows_grouped %>%
  mutate(across(c(ORIG_STUDY, DEST_STUDY), as.factor)) %>%
  gather_set_data(x = c("ORIG_GROUP", "DEST_GROUP"))

# Create the parallel sets plot with Study/Non-Study grouping
ggplot(flows_long, aes(x = x, id = id, split = y, value = DEPS)) +
  geom_parallel_sets(aes(fill = ORIG_GROUP), alpha = 0.6, axis.width = 0.1) +
  geom_parallel_sets_axes(axis.width = 0.1, fill = "gray90") +
  geom_parallel_sets_labels(angle = 0, size = 3, colour = "black") +
  scale_x_discrete(labels = c("Origin", "Destination")) +
  #scale_fill_brewer(type = "qual", palette = "Set2") +  # Use a color palette
  theme_minimal() +
  labs(
    title = "Flight Flows from Europe to Brazil (Study vs Non-Study)",
    subtitle = "Grouped by Origin and Destination Airport Categories",
    y = "Number of Departures"
  )

```

```{r}
ggplot(flows_long, aes(x = x, id = id, split = y, value = DEPS)) +
  geom_parallel_sets(aes(fill = ORIG_STUDY), alpha = 0.6, axis.width = 0.1) +
  geom_parallel_sets_axes(axis.width = 0.1, fill = "gray90") +
  geom_parallel_sets_labels(angle = 0, size = 3, colour = "black") +
  scale_x_discrete(labels = c("Origin", "Destination")) +
  scale_fill_manual(
    values = c("Study" = "blue", "Non-Study" = "lightblue")
    ) + # , 
  #                              "Study Brazil" = "red", "Non-Study Brazil" = "lightcoral")) +  # Custom color palette
  theme_minimal() +
  labs(
    title = "Flight Flows from Europe to Brazil (Study vs Non-Study)",
    subtitle = "Grouped by Origin and Destination Airport Categories",
    y = "Number of Departures"
  )

```

```{r}
ggplot(flows_long, aes(x = x, id = id, split = y, value = DEPS)) +
  geom_parallel_sets(aes(fill = ORIG_STUDY), alpha = 0.6, axis.width = 0.1) +
  geom_parallel_sets_axes(axis.width = 0.1, fill = "gray90") +
  geom_parallel_sets_labels(angle = 0, size = 3, colour = "black") +
  scale_x_discrete(labels = c("Origin", "Destination")) +
  scale_fill_manual(values = c("Study" = "blue", "Non-Study" = "lightblue")) +
  theme_minimal() +
  labs(
    # title = "Flight Flows from Europe to Brazil (Study vs Non-Study)",
    # subtitle = "Grouped by Origin and Destination Study Status",
    y = "Number of Departures", x = element_blank(), fill = element_blank()
  ) +
  theme(legend.position = "inside", legend.position.inside = c(0.1,0.1))

```


flows_grouped <- flows_grouped %>%
  mutate(
    ORIG_LABEL = case_when(
      ORIG_GROUP %in% study_airports_europe ~ paste0("Study: ", ORIG_GROUP),
      ORIG_GROUP %in% top_airports_europe ~ paste0("Non-Study: ", ORIG_GROUP),
      TRUE ~ paste0("Other: ", ORIG_GROUP)
    ),
    DEST_LABEL = case_when(
      DEST_GROUP %in% study_airports_brazil ~ paste0("Study: ", DEST_GROUP),
      DEST_GROUP %in% top_airports_brazil ~ paste0("Non-Study: ", DEST_GROUP),
      TRUE ~ paste0("Other: ", DEST_GROUP)
    )
  )

# Custom order of factor levels
all_airports <- unique(c(flows_grouped$ORIG_LABEL, flows_grouped$DEST_LABEL))

study_levels <- sort(unique(grep("^Study", all_airports, value = TRUE)))
non_study_levels <- sort(unique(grep("^Non-Study", all_airports, value = TRUE)))
other_levels <- sort(unique(grep("^Other", all_airports, value = TRUE)))

ordered_levels <- c(study_levels, non_study_levels, other_levels)

# Apply the factor levels
flows_grouped <- flows_grouped %>%
  mutate(
    ORIG_LABEL = factor(ORIG_LABEL, levels = ordered_levels),
    DEST_LABEL = factor(DEST_LABEL, levels = ordered_levels)
  )

flows_long <- flows_grouped %>%
  gather_set_data(x = c("ORIG_LABEL", "DEST_LABEL"))

ggplot(flows_long, aes(x = x, id = id, split = y, value = DEPS)) +
  geom_parallel_sets(aes(fill = ORIG_LABEL), alpha = 0.6, axis.width = 0.1) +
  geom_parallel_sets_axes(axis.width = 0.1, fill = "gray95") +
  geom_parallel_sets_labels(angle = 0, size = 3, colour = "black") +
  scale_x_discrete(labels = c("Origin", "Destination")) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(9, "Set1")[1:length(study_levels)]) +
  theme_minimal() +
  labs(
    title = "Flight Flows: Brazil–Europe",
    subtitle = "Airports grouped and ordered by Study, Non-Study, Other",
    y = "Departures"
  )


```{r}
# trim connections
eur_bra_conns <- eur_mvts |> 
  filter(DEP_PRU == "Eurocontrol", DES_CNTRY == "BR") |> 
  group_by(ADEP, ADES) |> reframe(DEPS = n()) |> arrange(desc(DEPS)) 

conns_threshold <- 90

major_nonstudy_eur <- eur_bra_conns |> 
  filter(!(ADEP %in%  study_eu)) |>   # only exclude study airports on the origin side
  group_by(ADEP)  |> 
  reframe(TOT_DEPS = sum(DEPS))  |> 
  filter(TOT_DEPS >= conns_threshold) %>%
  pull(ADEP)

major_nonstudy_bra <- eur_bra_conns |> 
  filter(!(ADES %in%  study_br)) |>   # only exclude study airports on the origin side
  group_by(ADES)  |> 
  reframe(TOT_DEPS = sum(DEPS))  |> 
  filter(TOT_DEPS >= conns_threshold) %>%
  pull(ADES)

eur_bra_grouped <- eur_bra_conns %>%
  mutate(
    ORIG_NAME = case_when(
      ADEP %in% study_eu ~ eu_labels[ADEP],
      ADEP %in% major_nonstudy_eur ~ ADEP,
      TRUE ~ "Other Europe"
    ),
    DEST_NAME = case_when(
      ADES %in% study_br ~ br_labels[ADES],
      ADES %in% major_nonstudy_bra ~ ADES,
      TRUE ~ "Other Brazil"
    )
  )

# Make sure both ORIG_GROUP and DEST_GROUP are factors (important!)
flows_grouped <- eur_bra_grouped %>%
  mutate(
    ORIG_GROUP = ifelse(ADEP %in% eur_apts, "Study", "Non-Study"),
    DEST_GROUP = ifelse(ADES %in% bra_apts,  "Study", "Non-Study")
  )

# prep for parallel sets plot
flows_long <- flows_grouped |> 
  # need to gather to feed set_parallel !!!!
  gather_set_data(x = c("ORIG_NAME","DEST_NAME")) |>
#add other stuff
mutate(GROUP_TYPE = case_when(
    ORIG_GROUP == "Study" ~ "blue"
  , DEST_GROUP == "Study" ~ "blue"
  , .default = "gray80"))

## Step 1: Get unique y/group_type mapping
# axis_fill_map <- flows_long %>%
#   select(y, GROUP_TYPE) %>%
#   distinct()

# Step 2: Join back to deduplicated mapping
# flows_long_fixed <- flows_long %>%
#   select(-GROUP_TYPE) %>%
#   left_join(axis_fill_map, by = "y")

flows_long |> mutate(YEAR = 2024) |> 
  write_csv("./data/BRA-EUR-conns-parallel.csv")
```

```{r}
flows_long <- read_csv(here::here("data","BRA-EUR-conns-parallel.csv"), show_col_types = FALSE)

ggplot(flows_long, aes(x = x, id = id, split = y, value = DEPS)) +
  geom_parallel_sets(aes(fill = ORIG_GROUP), alpha = 0.6, axis.width = 0.1) +
  geom_parallel_sets_axes(axis.width = 0.2, fill = "gray90") +
  geom_parallel_sets_labels(angle = 0, size = 3, colour = "black") +
  scale_x_discrete(labels = c("Origin", "Destination")) +
  scale_fill_manual(values = c("Study" = "blue", "Non-Study" = "lightblue")) +
  theme_minimal() +
  labs(
    # title = "Flight Flows from Europe to Brazil (Study vs Non-Study)",
    # subtitle = "Grouped by Origin and Destination Study Status",
    y = "Number of Departures", x = element_blank(), fill = element_blank()
  ) +
  theme(legend.position = "inside", legend.position.inside = c(0.1,0.1))
```

