---
title: "BRA-EUR - Towards Refining the Network Analysis"
format: pdf
---

```{r}
#| label: setup
#| message: false
#| warnin: false

source(here::here("_chapter-setup.R")) # all standard packages and defaults
```


# Overview

The idea is to develop a refined network analysis.    
For this we load the data for 2024 for Brazil (i.e. ANAC data set) and Europe (NM flight table).

```{r}
# -------------------- load the Brazilian network data ------------------------
cols <- c(
  "sg_empresa_icao",
  "nr_voo",
  "nm_mes_partida_real",
  "nr_mes_partida_real",
  "sg_icao_origem",
  "sg_icao_destino",
  "lt_combustivel",
  "sg_equipamento_icao",
  "nm_regiao_origem",
  "nm_pais_origem",
  "nm_regiao_destino",
  "nm_pais_destino",
  "nr_horas_voadas",
  NULL
)

bra_mvts <- read_csv("./data_proc/flights_ANAC_2024.csv"
                   #  , col_select = dplyr::any_of(cols)
                     , show_col_types = FALSE
                     ) |> 
  rename(
    ADEP = sg_icao_origem,
    ADES = sg_icao_destino,
    OPR  = sg_empresa_icao,            # aircraft operator (ICAO)
    FUEL_BURN = lt_combustivel,
    TYPE = sg_equipamento_icao,        # aircraft type (ICAO)
    REGIAO_DEP = nm_regiao_origem,
    DEP_CNTRY = nm_pais_origem,        # departing country
    REGIAO_ARR = nm_regiao_destino,
    DES_CNTRY = nm_pais_destino,       # destination country 
    FLT_DUR = nr_horas_voadas         # flight duration
)
```

```{r}
# ---------------- Europe --------------------------------------------------
# data is stored on local network drive
source(here::here("R","rqutils-zip.R"))
pth_to_data <- here::here(here::here() |> dirname(), "__DATA","NM-flight-table")
payload_zip <- "NM-flt-2024.zip"

check_zip_content(pth_to_data, payload_zip)  # check what we have in zip 

eur_mvts  <- read_zip(pth_to_data, payload_zip)
apts_meta <- arrow::read_parquet("./data/airport-icao-meta.parquet")

#| Utility function to prepare data -------------------------------------------
#| 
prep_eur_network_data <- function(.eur_mvts, .apts_meta = apts_meta){
  this_mvts <- .eur_mvts |> 
    dplyr::select(ADEP, ADES, FLTID = AIRCRAFT_ID, OPR = AIRCRAFT_OPERATOR
                  , REG = REGISTRATION, TYPE = AIRCRAFT_TYPE_ICAO_ID
                  , FLT_DUR = FLT_DUR_3
                  , DATE = LOBT
                  , SAM_ID = ID 
                  ) |> 
    dplyr::mutate(DATE = date(DATE))
  
  # augment airport details
  this_mvts <- this_mvts |> 
    dplyr::left_join(.apts_meta |> 
                       dplyr::select(ADEP = ICAO, DEP_CNTRY = COUNTRY_ISO)
                     , by = dplyr::join_by(ADEP)) |> 
    dplyr::left_join(.apts_meta |> 
                       dplyr::select(ADES = ICAO, DES_CNTRY = COUNTRY_ISO)
                     , by = dplyr::join_by(ADES))

  return(this_mvts)
}

eur_mvts <- eur_mvts |> dplyr::bind_rows() |> prep_eur_network_data()
  # all 2024 flights within EUR airspace ---------------------------
#  eur_mvts |> arrow::write_parquet(
#          "./data-prep-eur/EUR-network-connectivity-analytic-2024.parquet"
#)
  # write out study airport data -----------------------------------
# eur_mvts |> 
#   filter(ADEP %in% eur_apts | ADES %in% eur_apts) |> 
#   arrow::write_parquet(
#          "./data/EUR-network-connectivity-analytic-2024.parquet")
```




```{r}
calc_summary_routes_dep <- function(.mvts){
  this_summary <- .mvts |> 
    dplyr::group_by(ADEP) |> 
    dplyr::reframe(
       DEPS    = dplyr::n()
      ,NBR_DES = dplyr::n_distinct(ADES)) |> 
    dplyr::arrange(dplyr::desc(NBR_DES))
  
  return(this_summary)
}

bra_deps_routes <- bra_mvts |> calc_summary_routes_dep()
eur_deps_routes <- eur_mvts |> bind_rows() |> calc_summary_routes_dep()

# TODO - save as interim data and work from there
```



```{r}
bind_rows(
   bra_deps_routes |> filter(ADEP %in% bra_apts) |> mutate(REGION = "BRA")
  ,eur_deps_routes |> filter(ADEP %in% eur_apts) |> mutate(REGION = "EUR") 
  ) |> 
  mutate(ADEP = fct_infreq(ADEP, NBR_DES) |> fct_rev()) |> 
  ggplot() +
  geom_col(aes(x = NBR_DES, y = ADEP, fill = REGION)) +
  scale_fill_manual(values = bra_eur_colours) +
  facet_wrap(.~ REGION, scales = "free_y") +
  labs(y = element_blank()) +
  theme(legend.position = "none")
```

```{r}
get_spread_of_destinations <- function(.deps){
  this_spread <- .deps |> 
    dplyr::group_by(ADEP, DES_CNTRY) |> 
    dplyr::reframe(
            DEPS = dplyr::n()
          , DEST = dplyr::n_distinct(ADES)) |> 
    dplyr::group_by(ADEP) |> 
    dplyr::mutate(
           SHARE_DEP      = DEPS / sum(DEPS)
         , TOP_DEST_SHARE = DEST / sum(DEST)
         ) |> 
    dplyr::arrange(desc(DEPS), .by_group = TRUE) |> 
    dplyr::ungroup()
  
  return(this_spread)
}

#------------------- BRA ----------------------------------------------
bra_dests <- bra_mvts |> 
  filter(ADEP %in% bra_apts) |> 
  get_spread_of_destinations()

#------------------ EUR --------------------------------------------------
eur_mvts_study <- arrow::read_parquet("./data/EUR-network-connectivity-analytic-2024.parquet")
ectrl_ms       <- readxl::read_xlsx("./data/ECTRL-ICAOEUR-state-memberships.xlsx") |>   filter(ECTRL == 1) |> select(STATE_ISO = `State-ISO`, STATE = State) |> 
  mutate(REGION = "EUR")

#------------ STATE and ISO CODES ----------------------------------------
global_iso2    <- ISOcodes::ISO_3166_1 |> 
  tibble() |> 
  select(STATE_ISO = Alpha_2, STATE = Name)

patch_global <- tibble::tribble(
  ~ STATE_ISO, ~STATE,
  "AN", "Netherlands Antilles",
  "XK", "Kosovo"      # need to think how to add to EUR region
)
global_iso2 <- global_iso2 |> bind_rows(patch_global)

# world wide codes - we may need this for generalising BRA data
# constructed wiht ChatGPT - not validated !!!!
icao_iso2 <- read_csv("./data/country-icao-iso-territory.csv")

oa_url <- "https://raw.githubusercontent.com/davidmegginson/ourairports-data/refs/heads/main/airports.csv"
oa <- readr::read_csv(oa_url, show_col_types = F) 
oa <- oa |> 
  dplyr::filter(!is.na(icao_code)) |> 
  dplyr::select(ICAO = icao_code, NAME = name, CNTRY_ISO = iso_country)

# patch oa
patch_oa <- tibble::tribble(
  ~ ICAO, ~ NAME, ~ CNTRY_ISO
  ,"SBCO", "tbd" ,"BR"
  ,"SBJD", "tbd", "BR"    
  ,"SBJI", "tbd", "BR"
  ,"SBMI", "tbd", "BR"
 ,"SBMT" , "tbd", "BR"
 ,"SBPO", "tbd", "BR"
 ,"SBRD" , "tbd", "BR"
 ,"SBSG" , "tbd", "BR"
 ,"SBSI", "tbd", "BR"
 ,"SBSO", "tbd", "BR"
,"SBUY", "tbd", "BR"
,"SBVC" , "tbd", "BR"
,'GQNN', NA , "MR"    # Maritania
,'SDAG', NA ,'BR'
,'SDIY', NA ,'BR'
,'SDJV', NA ,'BR'
,'SDLO', NA ,'BR'
,'SDVZ', NA ,'BR'
,'SIFC', NA ,'BR'
,'SIRI', NA ,'BR'
,'SIUA', NA ,'BR'
,'SIXE', NA ,'BR'
,'SJZA', NA ,'BR'
,'SN6L', NA ,'BR'
,'SNAB', NA ,'BR'
,'SNBA', NA ,'BR'
,'SNCL', NA ,'BR'
,"SNEB", NA, "BR"
,"SNIG" , NA, "BR"
,"SNJM"  , NA, "BR"
,"SNLN" , NA, "BR"
,"SNRJ"  , NA, "BR"
,"SNSM"  , NA, "BR"
,"SNSS"  , NA, "BR"
,"SNTS"  , NA, "BR"
,"SNWS"  , NA, "BR"
,"SNZR"  , NA, "BR"
,"SSBN"  , NA, "BR"
,"SSGG"  , NA, "BR"
,"SSQO"  , NA, "BR"
,"SSRS"  , NA, "BR"
,"SSVL"  , NA, "BR"
,"SWBE"  , NA, "BR"
,"SWKQ"  , NA, "BR"
,"SWYN"  , NA, "BR"
,"SYEC"  , NA, "GN"
)
oa <- oa |> dplyr::bind_rows(patch_oa)

# icao_iso_other <- countrycode::codelist |> 
#   select(iso2c, country.name.en, continent, eurocontrol_statfor, eurocontrol_pru, icao.region)
# icao_iso_other |> readr::write_csv("./data/country-icao-iso-etc.csv")
icao_iso_other <- readr::read_csv("./data/country-icao-iso-etc.csv", show_col_types = F)

# to-do patch
#   iso2c country.name.en continent eurocontrol_statfor eurocontrol_pru icao.region
#   <chr> <chr>           <chr>     <chr>               <chr>           <chr>      
# 1 SN    Senegal         Africa    NA                  NA              AFI   


#------------------------------------------------------ END ISO ----------

eur_dests <- eur_mvts_study |> 
  filter(ADEP %in% eur_apts) |> 
  get_spread_of_destinations() |> 
  #---------- append ECTRL / EUR ---------
  left_join(global_iso2 |> 
              rename(DES_CNTRY = STATE_ISO), by = join_by(DES_CNTRY)
            ) |>   
  left_join(ectrl_ms |> select(-STATE) |> 
              rename(DES_CNTRY = STATE_ISO) , by = join_by(DES_CNTRY)
            )
```

```{r}
p <- bra_dests |> 
  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines")
  ) 
p
p + theme(legend.position = "none")
```
international traffic

```{r}
bra_dests |> filter(DES_CNTRY != "BRASIL") |> 
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev())
```


```{r}
# DES_CNTRY ~ PAIS_DESTINO

bra_dests |> filter(DES_CNTRY != "BRASIL") |> 
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |> 
  
  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(fill = element_blank()) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines"),
    legend.position = "inside", legend.position.inside = c(0.75,0.25)
  ) 
```

```{r}
eur_dests |>
  filter(is.na(REGION)) |>
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |>

  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(fill = element_blank()) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines"),
    legend.position = "inside", legend.position.inside = c(0.75,0.25)
  )
```

Appendix 1 to the Directives to Regional Air Navigation Meetings and Rules of Procedure for their Conduct (Doc 8144-AN/874), comprises the following regions:

* Africa-Indian Ocean Region (AFI);
* Asia Region (ASIA);
* Caribbean Region (CAR);
* European Region (EUR);
* Middle East Region (MID);
* North American Region (NAM);
* North Atlantic Region (NAT);
* Pacific Region (PAC);
* South American Region (SAM).

```{r}
eur_world_shares <- eur_dests |>
  filter(is.na(REGION)) |>
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |> 
  left_join(icao_iso2 |> rename(DES_CNTRY = `ISO Alpha-2 Code`, ICAO_REG = `ICAO Region`) |> 
              select(DES_CNTRY, ICAO_REG)) |> 
  # aggregate by ICAO Region
  group_by(ICAO_REG) |> 
  reframe(DEPS = sum(DEPS), DEST = sum(DEST)) |> 
  mutate(REG_SHARES = DEPS / sum(DEPS))

eur_world_shares
```
```{r}
eur_world_shares |> 
  ggplot() +
  geom_col(aes(x = REG_SHARES, y = ICAO_REG, fill = ICAO_REG))
```

EUR ~  non-ECTRL Europe (probably need to take this out)



# Workflow Demo - Getting from Code to Functions

```{r}
calc_adep_counts <- function(.ds){
  my_count <- .ds |> 
  group_by(ADEP, DEP_CNTRY) |>  # Group by ADEP and PAIS_ORIGEM
  summarise(count = n(), .groups = "drop") |>  
  arrange(desc(count))
  
  return(my_count)
}



bra_mvts |> calc_adep_counts() 

plot_barchart <- function(.ds){
  
  p <- .ds |> 
  ggplot() +
  geom_col(aes(x = ADEP, y = count)) +
    facet_wrap(.~ REGION)
  
  return(p)
}

bra_demo <- bra_mvts |> calc_adep_counts() |> mutate(REGION = "BRA")
eur_demo <- bra_mvts |> calc_adep_counts() |> mutate(REGION = "EUR")

p1 <- bra_demo |> plot_barchart()
p2 <- eur_demo |> plot_barchart()

p1 / p2

bind_rows(bra_demo, eur_demo) |> plot_barchart()
```

The maximum departures seen at Brazilian airport account for `r bra_demo |> pull(count) |> max()`.





# Prepare analytic data for study

[1] "ADEP"      "ADES"      "FLTID"     "OPR"       "REG"       "TYPE"      "FLT_DUR"  
 [8] "DATE"      "SAM_ID"    "DEP_CNTRY" "DES_CNTRY"

```{r}
bra_mvts <- read_csv("./data_proc/flights_ANAC_2024.csv"
                   #  , col_select = dplyr::any_of(cols)
                     , show_col_types = FALSE
                     ) 
trim_ANAC <- function(.anac){
  trimmed_anac <- .anac |> 
    dplyr::select(
       ADEP    = sg_icao_origem
      ,ADES    = sg_icao_destino
      ,FLT_NBR = nr_voo
      ,OPR  = sg_empresa_icao           # aircraft operator (ICAO)
      ,REG  = ds_matricula 
      ,TYPE = sg_equipamento_icao       # aircraft type (ICAO)
      ,AOBT    = hr_partida_real
      ,AIBT    = hr_chegada_real
      ,FLT_DUR = nr_horas_voadas        # flight duration
      ,DATE    = dt_referencia
      ,id_basica                        # ANAC ID
      ,DEP_CNTRY_NAME = nm_pais_origem  # departing country
      ,DES_CNTRY_NAME = nm_pais_destino # destination country
    )
  return(trimmed_anac)
}

prep_bra_connectivity <- function(.trimmed_anac){
  bra_analytic <- .trimmed_anac |> 
    dplyr::mutate(
       FLTID = paste0(OPR, FLT_NBR)
    )
 return(bra_analytic)
}

append_iso_adep_ades <- function(
      .flt_adep_ades
    , .iso_look_up = oa |> select(ICAO, CNTRY_ISO)
    ){
  this_flt <- .flt_adep_ades |> 
    dplyr::left_join(
       .iso_look_up |> dplyr::rename(ADEP = ICAO, DEP_CNTRY = CNTRY_ISO)
      , by = dplyr::join_by(ADEP)
      ) |> 
    dplyr::left_join(
      .iso_look_up |> dplyr::rename(ADES = ICAO, DES_CNTRY = CNTRY_ISO)
      , by = dplyr::join_by(ADES))
  
  return(this_flt)
}

append_ades_regions <- function(
     .flt_ades_iso2
    ,.region_lookup = icao_iso_other |> 
      filter(!is.na(iso2c)) |> 
      select(  DES_CNTRY = iso2c
             , DES_ICAO  = icao.region   
             , DES_STATF = eurocontrol_statfor
             , DES_PRU = eurocontrol_pru)
){
  this_flt <- .flt_ades_iso2 |> 
    dplyr::left_join(.region_lookup, by = dplyr::join_by(DES_CNTRY))
  
  return(this_flt)
}

bra_mvts <- bra_mvts |> trim_ANAC() |> append_iso_adep_ades() |> append_ades_regions()
```


```{r}
bra_reg_tfc <- bra_mvts |> 
  mutate(DOMESTIC = DES_CNTRY == "BR") |> 
  group_by(DES_ICAO, DOMESTIC) |> 
  reframe(DEPS = n())

bra_reg_tfc |> mutate(LABEL = ifelse(!DOMESTIC, DES_ICAO, "local"), SHARE = DEPS / sum(DEPS)) |> mutate(LABEL = forcats::fct_reorder(LABEL, SHARE)) |> drop_na() |> ggplot() + geom_col(aes(x = SHARE, y = LABEL, fill = LABEL)) + scale_x_continuous(labels = scales::percent_format()) + labs(x = element_blank(), y = element_blank())
```


```{r}
library(ComplexUpset)

dests_bra_eur <- bra_mvts |> 
  filter(DES_PRU == "Eurocontrol", DEP_CNTRY == "BR") |> 
  group_by(ADEP, ADES) |> 
  reframe(DEPS = n()) |> 
  arrange(desc(DEPS)) |> 
  mutate(SHARE = DEPS / sum(DEPS), DUMMY = 1
         ,CUM_SHARE = cumsum(SHARE)) 

tmp <- bra_mvts |> 
  filter(DEP_CNTRY == "BR", DES_PRU == "Eurocontrol") |> 
  group_by(ADEP, ADES, DES_PRU, DEP_CNTRY) |> 
  reframe(DEPS = n()) |> 
  arrange(desc(DEPS)) |> 
  mutate(SHARE = DEPS / sum(DEPS), CUM_SHARE = cumsum(SHARE)) |> 
  mutate(IS_STUDY = ADES %in% eur_apts) |> 
  pivot_wider(names_from = ADES, values_from = IS_STUDY)

tmp <- tmp |> filter(CUM_SHARE <= 0.8)
tiks <- names(tmp)[-c(1:4)]
tmp |> upset(tiks)
```


another cool and long airport lookup

```{r}
ip2_url <- "https://github.com/ip2location/ip2location-iata-icao/blob/master/iata-icao.csv"

# github differentiates between "display" and "raw" version
# we download raw versions!
ip2_url <- "https://raw.githubusercontent.com/ip2location/ip2location-iata-icao/refs/heads/master/iata-icao.csv"


icao_ip2 <- readr::read_csv(ip2_url)
```

