---
title: "BRA-EUR - Towards Refining the Network Analysis"
format: pdf
---

```{r}
#| label: setup
#| message: false
#| warnin: false

source(here::here("_chapter-setup.R")) # all standard packages and defaults
```


# Overview

The idea is to develop a refined network analysis.    
For this we load the data for 2024 for Brazil (i.e. ANAC data set) and Europe (NM flight table).

```{r}
# -------------------- load the Brazilian network data ------------------------
cols <- c(
  "sg_empresa_icao",
  "nr_voo",
  "nm_mes_partida_real",
  "nr_mes_partida_real",
  "sg_icao_origem",
  "sg_icao_destino",
  "lt_combustivel",
  "sg_equipamento_icao",
  "nm_regiao_origem",
  "nm_pais_origem",
  "nm_regiao_destino",
  "nm_pais_destino",
  "nr_horas_voadas",
  NULL
)

bra_mvts <- read_csv("./data_proc/flights_ANAC_2024.csv"
                     , col_select = dplyr::any_of(cols)
                     , show_col_types = FALSE
                     ) |> 
  rename(
    ADEP = sg_icao_origem,
    ADES = sg_icao_destino,
    OPR  = sg_empresa_icao,            # aircraft operator (ICAO)
    FUEL_BURN = lt_combustivel,
    TYPE = sg_equipamento_icao,        # aircraft type (ICAO)
    REGIAO_DEP = nm_regiao_origem,
    DEP_CNTRY = nm_pais_origem,        # departing country
    REGIAO_ARR = nm_regiao_destino,
    DES_CNTRY = nm_pais_destino,       # destination country 
    FLT_DUR = nr_horas_voadas         # flight duration
)
```

```{r}
# ---------------- Europe --------------------------------------------------
# data is stored on local network drive
source(here::here("R","rqutils-zip.R"))
pth_to_data <- here::here(here::here() |> dirname(), "__DATA","NM-flight-table")
payload_zip <- "NM-flt-2024.zip"

check_zip_content(pth_to_data, payload_zip)  # check what we have in zip 

eur_mvts  <- read_zip(pth_to_data, payload_zip)
apts_meta <- arrow::read_parquet("./data/airport-icao-meta.parquet")

#| Utility function to prepare data -------------------------------------------
#| 
prep_eur_network_data <- function(.eur_mvts, .apts_meta = apts_meta){
  this_mvts <- .eur_mvts |> 
    dplyr::select(ADEP, ADES, FLTID = AIRCRAFT_ID, OPR = AIRCRAFT_OPERATOR
                  , REG = REGISTRATION, TYPE = AIRCRAFT_TYPE_ICAO_ID
                  , FLT_DUR = FLT_DUR_3
                  , DATE = LOBT
                  , SAM_ID = ID 
                  ) |> 
    dplyr::mutate(DATE = date(DATE))
  
  # augment airport details
  this_mvts <- this_mvts |> 
    dplyr::left_join(.apts_meta |> 
                       dplyr::select(ADEP = ICAO, DEP_CNTRY = COUNTRY_ISO)
                     , by = dplyr::join_by(ADEP)) |> 
    dplyr::left_join(.apts_meta |> 
                       dplyr::select(ADES = ICAO, DES_CNTRY = COUNTRY_ISO)
                     , by = dplyr::join_by(ADES))

  return(this_mvts)
}

eur_mvts <- eur_mvts |> dplyr::bind_rows() |> prep_eur_network_data()
  # all 2024 flights within EUR airspace ---------------------------
#  eur_mvts |> arrow::write_parquet(
#          "./data-prep-eur/EUR-network-connectivity-analytic-2024.parquet"
#)
  # write out study airport data -----------------------------------
# eur_mvts |> 
#   filter(ADEP %in% eur_apts | ADES %in% eur_apts) |> 
#   arrow::write_parquet(
#          "./data/EUR-network-connectivity-analytic-2024.parquet")
```




```{r}
calc_summary_routes_dep <- function(.mvts){
  this_summary <- .mvts |> 
    dplyr::group_by(ADEP) |> 
    dplyr::reframe(
       DEPS    = dplyr::n()
      ,NBR_DES = dplyr::n_distinct(ADES)) |> 
    dplyr::arrange(dplyr::desc(NBR_DES))
  
  return(this_summary)
}

bra_deps_routes <- bra_mvts |> calc_summary_routes_dep()
eur_deps_routes <- eur_mvts |> bind_rows() |> calc_summary_routes_dep()

# TODO - save as interim data and work from there
```



```{r}
bind_rows(
   bra_deps_routes |> filter(ADEP %in% bra_apts) |> mutate(REGION = "BRA")
  ,eur_deps_routes |> filter(ADEP %in% eur_apts) |> mutate(REGION = "EUR") 
  ) |> 
  mutate(ADEP = fct_infreq(ADEP, NBR_DES) |> fct_rev()) |> 
  ggplot() +
  geom_col(aes(x = NBR_DES, y = ADEP, fill = REGION)) +
  scale_fill_manual(values = bra_eur_colours) +
  facet_wrap(.~ REGION, scales = "free_y") +
  labs(y = element_blank()) +
  theme(legend.position = "none")
```

```{r}
get_spread_of_destinations <- function(.deps){
  this_spread <- .deps |> 
    dplyr::group_by(ADEP, DES_CNTRY) |> 
    dplyr::reframe(
            DEPS = dplyr::n()
          , DEST = dplyr::n_distinct(ADES)) |> 
    dplyr::group_by(ADEP) |> 
    dplyr::mutate(
           SHARE_DEP      = DEPS / sum(DEPS)
         , TOP_DEST_SHARE = DEST / sum(DEST)
         ) |> 
    dplyr::arrange(desc(DEPS), .by_group = TRUE) |> 
    dplyr::ungroup()
  
  return(this_spread)
}

#------------------- BRA ----------------------------------------------
bra_dests <- bra_mvts |> 
  filter(ADEP %in% bra_apts) |> 
  get_spread_of_destinations()

#------------------ EUR --------------------------------------------------
eur_mvts_study <- arrow::read_parquet("./data/EUR-network-connectivity-analytic-2024.parquet")
ectrl_ms       <- readxl::read_xlsx("./data/ECTRL-ICAOEUR-state-memberships.xlsx") |>   filter(ECTRL == 1) |> select(STATE_ISO = `State-ISO`, STATE = State) |> 
  mutate(REGION = "EUR")

#------------ STATE and ISO CODES ----------------------------------------
global_iso2    <- ISOcodes::ISO_3166_1 |> 
  tibble() |> 
  select(STATE_ISO = Alpha_2, STATE = Name)

patch_global <- tibble::tribble(
  ~ STATE_ISO, ~STATE,
  "AN", "Netherlands Antilles",
  "XK", "Kosovo"      # need to think how to add to EUR region
)
global_iso2 <- global_iso2 |> bind_rows(patch_global)

# world wide codes - we may need this for generalising BRA data
# constructed wiht ChatGPT - not validated !!!!
icao_iso2 <- read_csv("./data/country-icao-iso-territory.csv")

#------------------------------------------------------ END ISO ----------

eur_dests <- eur_mvts_study |> 
  filter(ADEP %in% eur_apts) |> 
  get_spread_of_destinations() |> 
  #---------- append ECTRL / EUR ---------
  left_join(global_iso2 |> 
              rename(DES_CNTRY = STATE_ISO), by = join_by(DES_CNTRY)
            ) |>   
  left_join(ectrl_ms |> select(-STATE) |> 
              rename(DES_CNTRY = STATE_ISO) , by = join_by(DES_CNTRY)
            )
```

```{r}
p <- bra_dests |> 
  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines")
  ) 
p
p + theme(legend.position = "none")
```
international traffic

```{r}
bra_dests |> filter(DES_CNTRY != "BRASIL") |> 
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev())
```


```{r}
# DES_CNTRY ~ PAIS_DESTINO

bra_dests |> filter(DES_CNTRY != "BRASIL") |> 
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |> 
  
  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(fill = element_blank()) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines"),
    legend.position = "inside", legend.position.inside = c(0.75,0.25)
  ) 
```

```{r}
eur_dests |>
  filter(is.na(REGION)) |>
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |>

  ggplot() + geom_col(aes(x = SHARE_DEP, y = ADEP, fill = DES_CNTRY)) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(fill = element_blank()) +
  theme(
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.3, "lines"),
    legend.position = "inside", legend.position.inside = c(0.75,0.25)
  )
```

Appendix 1 to the Directives to Regional Air Navigation Meetings and Rules of Procedure for their Conduct (Doc 8144-AN/874), comprises the following regions:

* Africa-Indian Ocean Region (AFI);
* Asia Region (ASIA);
* Caribbean Region (CAR);
* European Region (EUR);
* Middle East Region (MID);
* North American Region (NAM);
* North Atlantic Region (NAT);
* Pacific Region (PAC);
* South American Region (SAM).

```{r}
eur_world_shares <- eur_dests |>
  filter(is.na(REGION)) |>
  mutate(ADEP = fct_infreq(ADEP, SHARE_DEP) |> fct_rev()) |> 
  left_join(icao_iso2 |> rename(DES_CNTRY = `ISO Alpha-2 Code`, ICAO_REG = `ICAO Region`) |> 
              select(DES_CNTRY, ICAO_REG)) |> 
  # aggregate by ICAO Region
  group_by(ICAO_REG) |> 
  reframe(DEPS = sum(DEPS), DEST = sum(DEST)) |> 
  mutate(REG_SHARES = DEPS / sum(DEPS))

eur_world_shares
```
```{r}
eur_world_shares |> 
  ggplot() +
  geom_col(aes(x = REG_SHARES, y = ICAO_REG, fill = ICAO_REG))
```

EUR ~  non-ECTRL Europe (probably need to take this out)



# Workflow Demo - Getting from Code to Functions

```{r}
calc_adep_counts <- function(.ds){
  my_count <- .ds |> 
  group_by(ADEP, DEP_CNTRY) |>  # Group by ADEP and PAIS_ORIGEM
  summarise(count = n(), .groups = "drop") |>  
  arrange(desc(count))
  
  return(my_count)
}



bra_mvts |> calc_adep_counts() 

plot_barchart <- function(.ds){
  
  p <- .ds |> 
  ggplot() +
  geom_col(aes(x = ADEP, y = count)) +
    facet_wrap(.~ REGION)
  
  return(p)
}

bra_demo <- bra_mvts |> calc_adep_counts() |> mutate(REGION = "BRA")
eur_demo <- bra_mvts |> calc_adep_counts() |> mutate(REGION = "EUR")

p1 <- bra_demo |> plot_barchart()
p2 <- eur_demo |> plot_barchart()

p1 / p2

bind_rows(bra_demo, eur_demo) |> plot_barchart()
```

The maximum departures seen at Brazilian airport account for `r bra_demo |> pull(count) |> max()`.

