---
title: "Lisbon Airport"
format: typst
---

This Quarto document lists the data preparatory action for integrating Lisbon airport, LPPT.
The APDF data for LPPT was extracted from ECTRL data warehouse and stored locally.    
This script is not reproducible if the ADPF data are not accessible.

```{r}
#| label: setup

library(tidyverse)

source(here::here("R","rqutils-zip.R"))
source(here::here("R","rq-apdf-to-punc.R"))
source(here::here("R","rq-apdf-to-pbwg.R"))     # todo- clean - need to load punc first.
source(here::here("03-traffic_characterization-graphs.R"))
```

```{r}
#| label: read-LPPT-zip

lppt_zip    <- list.files(pth_apdf, pattern = "^LPPT-apdf")
whats_there <- check_zip_content(pth_apdf, lppt_zip)
whats_there
```

## Airport Traffic

For BRA-EUR, we use PBWG daily traffic stats as analytic data.

```{r}
# let's iterate over each year
whats_there <- whats_there |> mutate(YEAR = str_sub(Name, 13,16))

# loop for one
read_one_year <- function(.pth_apdf, .archive, .yr, ...){
  this_target <- whats_there |> dplyr::filter(YEAR == .yr)
  this_data   <- read_zip(.pth_apdf, .archive, .files = this_target$Name)
 #  this_data |> read_and_write_apt_tfc("LPPT", .yr)
}

# test-run: read_one_year(pth_apdf, lppt_zip, .yr = 2019) |> print()
# do all
whats_there$YEAR |> purrr::walk(.f = ~ read_one_year(pth_apdf, lppt_zip, .x))
```

* document write out of traffic (count) data set
* script changed to 


Check what we have - traffic data set

```{r}
tfc_data <- list.files(here::here("data-prep-eur"), pattern = "apt-tfc-LPPT", full.names = T) |> 
  purrr::map(.f = ~ readr::read_csv(.x, show_col_types = F))
```

```{r}
lppt_annual_tfc <- tfc_data |> bind_rows() |> 
  mutate(YEAR = lubridate::year(DATE)) |> 
  group_by(ICAO, YEAR) |> 
  reframe(MVTS = sum(ARRS) + sum(DEPS))

lppt_annual_tfc |> gt::gt()
```
```{r}
tfc_apts_bra <- read_csv("./data/BRA-airport-tfc.csv", show_col_types = FALSE)
sbgr_tfc_annual <- tfc_apts_bra |> filter(ICAO %in% "SBGR") |> mutate(YEAR = lubridate::year(DATE)) |> group_by(ICAO, YEAR) |> reframe(MVTS = sum(ARRS) + sum(DEPS))

lppt_annual_tfc |> ggplot() + 
  geom_col(aes(y = MVTS, x = YEAR)) +
  geom_line(data = sbgr_tfc_annual,aes(y = MVTS, x = YEAR, group = ICAO), color = "green" ) +
  geom_text( 
     data = sbgr_tfc_annual |> filter(YEAR == 2024)
    ,aes(y = MVTS, x = 2024, label = ICAO)
    , color = "green" 
    , vjust = 1.2 )
```

## Airport Peak Traffic

based on BRA-EUR, i.e. 99th percentile

```{r}
lppt_annual_peak <- tfc_data |> bind_rows() |> peak_day_from_counts() |> 
    add_nbr_rwy() |> 
    mutate(REGION = "EUR")

lppt_annual_peak |> 
  ggplot(aes(x = YEAR, y = PEAK_DAY_PCT)) +
  geom_line(aes(group = ICAO)) +
  geom_point()
```


## Airport Punctuality

```{r}
lppt_adpf <- read_zip(pth_apdf, lppt_zip)

# do it for one
calc_punc_for_df_with_delay <- function(.df){
  this_punc <- .df |> 
  prep_apdf() |> 
  distinct() |> 
  mutate(ICAO = "LPPT") |> 
  add_delay_and_dlygrp() |> 
  package_pbwg_punc_analytic()
  return(this_punc)
}

punc_2019 <- lppt_adpf[[1]] |> calc_punc_for_df_with_delay()

# iterate over all
punc_lppt <-  lppt_adpf |> 
  purrr::map(.f = ~ .x |> calc_punc_for_df_with_delay())

punc_lppt |> bind_rows() |> 
  readr::write_csv(here::here("data","PBWG-EUR-PUNC-LPPT-2019-2024.csv"))
```

