---
title: "Lisbon Airport"
format: typst
---

This Quarto document lists the data preparatory action for integrating Lisbon airport, LPPT.
The APDF data for LPPT was extracted from ECTRL data warehouse and stored locally.    
This script is not reproducible if the ADPF data are not accessible.

```{r}
#| label: setup

library(tidyverse)

source(here::here("R","rqutils-zip.R"))
source(here::here("R","rq-apdf-to-punc.R"))
source(here::here("R","rq-apdf-to-pbwg.R"))     # todo- clean - need to load punc first.
source(here::here("R","rq-apdf-asma-dsma.R"))
source(here::here("R","03-traffic_characterization-graphs.R"))
```

```{r}
#| label: read-LPPT-zip

lppt_zip    <- list.files(pth_apdf, pattern = "^LPPT-apdf")
whats_there <- check_zip_content(pth_apdf, lppt_zip)
whats_there
```

## Airport Traffic

For BRA-EUR, we use PBWG daily traffic stats as analytic data.

```{r}
# let's iterate over each year
whats_there <- whats_there |> mutate(YEAR = str_sub(Name, 13,16))

# loop for one
read_one_year <- function(.pth_apdf, .archive, .yr, ...){
  this_target <- whats_there |> dplyr::filter(YEAR == .yr)
  this_data   <- read_zip(.pth_apdf, .archive, .files = this_target$Name)
 #  this_data |> read_and_write_apt_tfc("LPPT", .yr)
}

# test-run: read_one_year(pth_apdf, lppt_zip, .yr = 2019) |> print()
# do all
whats_there$YEAR |> purrr::walk(.f = ~ read_one_year(pth_apdf, lppt_zip, .x))
```

* document write out of traffic (count) data set
* script changed to 


Check what we have - traffic data set

```{r}
tfc_data <- list.files(here::here("data-prep-eur"), pattern = "apt-tfc-LPPT", full.names = T) |> 
  purrr::map(.f = ~ readr::read_csv(.x, show_col_types = F))
```

```{r}
lppt_annual_tfc <- tfc_data |> bind_rows() |> 
  mutate(YEAR = lubridate::year(DATE)) |> 
  group_by(ICAO, YEAR) |> 
  reframe(MVTS = sum(ARRS) + sum(DEPS))

lppt_annual_tfc |> gt::gt()
```
```{r}
tfc_apts_bra <- read_csv("./data/BRA-airport-tfc.csv", show_col_types = FALSE)
sbgr_tfc_annual <- tfc_apts_bra |> filter(ICAO %in% "SBGR") |> mutate(YEAR = lubridate::year(DATE)) |> group_by(ICAO, YEAR) |> reframe(MVTS = sum(ARRS) + sum(DEPS))

lppt_annual_tfc |> ggplot() + 
  geom_col(aes(y = MVTS, x = YEAR)) +
  geom_line(data = sbgr_tfc_annual,aes(y = MVTS, x = YEAR, group = ICAO), color = "green" ) +
  geom_text( 
     data = sbgr_tfc_annual |> filter(YEAR == 2024)
    ,aes(y = MVTS, x = 2024, label = ICAO)
    , color = "green" 
    , vjust = 1.2 )
```

## Airport Peak Traffic

based on BRA-EUR, i.e. 99th percentile   
the peak day (99th percentile) is calculated from the daily traffic stats

```{r}
lppt_annual_peak <- tfc_data |> bind_rows() |> peak_day_from_counts() |> 
    add_nbr_rwy() |> 
    mutate(REGION = "EUR")

lppt_annual_peak |> 
  ggplot(aes(x = YEAR, y = PEAK_DAY_PCT)) +
  geom_line(aes(group = ICAO)) +
  geom_point()
```


## Airport Punctuality

```{r}
lppt_adpf <- read_zip(pth_apdf, lppt_zip)

# do it for one
calc_punc_for_df_with_delay <- function(.df){
  this_punc <- .df |> 
  prep_apdf() |> 
  distinct() |> 
  mutate(ICAO = "LPPT") |> 
  add_delay_and_dlygrp() |> 
  package_pbwg_punc_analytic()
  return(this_punc)
}

punc_2019 <- lppt_adpf[[1]] |> calc_punc_for_df_with_delay()

# iterate over all
punc_lppt <-  lppt_adpf |> 
  purrr::map(.f = ~ .x |> calc_punc_for_df_with_delay())

# write out --- punctuality data for LPPT
# punc_lppt |> bind_rows() |> 
#  readr::write_csv(here::here("data","PBWG-EUR-PUNC-LPPT-2019-2024.csv"))
```

check what we have

```{r}
#| message: false
#| warning: false
#| echo: false

# load predictability helpers
source(here::here("R","04-predictability_new-graphs.R"))

# read punc data for LPPT 
punc_lppt <- read_csv(here::here("data","PBWG-EUR-PUNC-LPPT-2019-2024.csv"), show_col_types = F)


punc_lppt |> rename(APT = ICAO) |>  
    mutate(REGION = "EUR"
           , YEAR=lubridate::year(DATE)
           , MONTH=lubridate::month(DATE)
           ) |> 
    group_by(APT, PHASE, YEAR, MONTH, REGION) |>  # , N_VALID) |> 
    reframe(across(.cols = N_VALID:`[60,INF)`, .fns = sum), .groups = "drop") |> 
    add_dly_early_late_groupings2() |> 
  
    mutate(YM = paste0(YEAR,"-", sprintf("%02d", MONTH),"-01") |> 
             lubridate::ymd() |> lubridate::floor_date(unit = "month")
           , .before = PHASE) |> 
  
  ggplot() + geom_path(aes(x = YM, y = WITHIN_M05P05, group = APT, color = PHASE ))
```

:::{.callout-note}
CHECK FOR DATA IN 2nd Half of 2024

Data file has full 2024 -- check data types -- why does function fail?
:::

## Additional Taxi-Time

Per PBWG convention, we determine the reference times for the year 2024.

```{r}
source(here::here("R","rq-apdf-to-txxt.R"))
```


```{r}
lppt_2024 <- read_zip(pth_apdf, lppt_zip, "LPPT_01-JAN-2024_31-DEC-2024.gz.parquet")
txxt_ref_2024 <- lppt_2024 |> 
  rename(PHASE = SRC_PHASE, RWY = AP_C_RWY, STND = AP_C_STND) |>  
  prepare_eur_txxts() |> 
  calc_pct20_cases(ymd_hms("2024-01-01 00:00:00"), ymd_hms("2024-12-31 24:00:00")) |> 
  rename(N_2024 = N, REF_20_2024 = REF_20)

# compare with 2019
lppt_2019 <- read_zip(pth_apdf, lppt_zip, "LPPT_01-JAN-2019_31-DEC-2019.gz.parquet")
txxt_ref_2019 <- lppt_2019 |> 
  rename(PHASE = SRC_PHASE, RWY = AP_C_RWY, STND = AP_C_STND) |>  
  prepare_eur_txxts() |> 
  calc_pct20_cases(ymd_hms("2019-01-01 00:00:00"), ymd_hms("2019-12-31 24:00:00")) |> 
  rename(N_2019 = N, REF_20_2019 = REF_20)
# correct for runway change
txxt_ref_2019 <- txxt_ref_2019 |> 
  mutate( RWY_OLD = RWY
         ,RWY     = case_when(
                RWY %in% "03" ~ "02"
              , RWY %in% "21" ~ "20"
           )
         )
```

txxt_ref_2024 |> full_join(txxt_ref_2019, by = join_by(PHASE, RWY, STND)) |> 
  # mutate(DELTA = REF_20_2024 - REF_20_2019) |> write_csv(here::here("data-prep-eur","apt-txxt-ref-LPPT.csv"))

```{r}
# load reference taxi times for LPPT
lppt_txxt_refs <- read_csv(
      here::here("data-prep-eur","apt-txxt-ref-LPPT.csv")
    , show_col_types = FALSE) |> 
  # force ref time for all cases ----------------------------
  mutate(REF_20_2024 = case_when(
      is.na(REF_20_2024) & !is.na(REF_20_2019) ~ REF_20_2019
    , .default = REF_20_2024)
    ) |> 
  select(PHASE, RWY, STND, REF = REF_20_2024)
```

```{r}
# load all movement data 2019-2024
lppt_txxts <- read_zip(pth_apdf, lppt_zip) |> bind_rows() |> 
  rename(PHASE = SRC_PHASE, RWY = AP_C_RWY, STND = AP_C_STND) |> 
  mutate(ICAO = "LPPT", .before = PHASE) |> 
  prepare_eur_txxts() |> 
  # account for runway change
  mutate(RWY = case_when(
                RWY %in% "03" ~ "02"
              , RWY %in% "21" ~ "20"
           )
  )
```

```{r}
lppt_txxts <- lppt_txxts |> 
  left_join(lppt_txxt_refs, by = join_by(PHASE, RWY, STND)) |> 
  mutate(ADD_TXXT = TXXT - REF)

lppt_txxt_add_times <- lppt_txxts |> 
  mutate(ICAO = "LPPT", DATE = date(BLOCK_TIME_UTC)) |> 
  group_by(ICAO, DATE, PHASE) |> 
  reframe( MVTS     = sum(!is.na(ADD_TXXT))
          ,TXXT     = sum(TXXT, na.rm = TRUE)
          ,ADD_TIME = sum(ADD_TXXT, na.rm = TRUE)
          ) |> 
  mutate(AVG_ADD_TIME = ADD_TIME / MVTS)
```

lppt_txxt_add_times |> mutate(PHASE = ifelse(PHASE == "ARR", "TXIT", "TXOT")) |> 
  write_csv(here::here("data","EUR-txxt-LPPT.csv"))

```{r}
lppt_txxt_add_times |> 
  ggplot() + 
  geom_path(aes(x = DATE, y = AVG_ADD_TIME, color = PHASE, group = PHASE)) +
  facet_wrap(.~ PHASE) +
  labs(subtitle = "daily variation additional taxi-times") +
  theme(legend.position = "none")
```

## Additional Time in Terminal Airspace

BRA-EUR focusses on ASMA (KPI08) - additional time in the arrival airspace.

```{r}
# load the data sets
if(!exists("lppt_adpf")){
  lppt_adpf <- read_zip(pth_apdf, lppt_zip)
}

asma_dsma <- lppt_adpf |> bind_rows() |> extract_asma_dsma() |> prepare_eur_asma_dsma() |> mutate(DATE = lubridate::date(MVT_TIME))
```

ASMA sectors provided by Thierry:

```{r}
lppt_sects <- tibble::tribble(
~AIRPORT,	~ASMA_RADIUS,	~VERSION,	~VALID_FROM, ~VALID_TO,	~ASMA_SECTOR, ~FROM_BEARING,	~TO_BEARING
,"LPPT",	40,	0,	"01-Jan-11",	"01-Apr-24",	1,	40	,100
,"LPPT",	40,	0,	"01-Jan-11",	"01-Apr-24",	2,	100	,160
,"LPPT",	40,	0,	"01-Jan-11",	"01-Apr-24",	3,	160	,235
,"LPPT",	40,	0,	"01-Jan-11",	"01-Apr-24",	4,	235	,295
,"LPPT",	40,	0,	"01-Jan-11",	"01-Apr-24",	5,	295	,350
,"LPPT",	40,	0,	"01-Jan-11",	"01-Apr-24",	6,	350	,40
,"LPPT",	40,	1,	"01-Apr-24",	"01-Dec-99",	1,	40	,100
,"LPPT",	40,	1,	"01-Apr-24",	"01-Dec-99",	2,	100	,150
,"LPPT",	40,	1,	"01-Apr-24",	"01-Dec-99",	3,	150	,240
,"LPPT",	40,	1,	"01-Apr-24",	"01-Dec-99", 	4,	240	,300
,"LPPT",	40,	1,	"01-Apr-24",  "01-Dec-99",	5,	300	,350
,"LPPT",	40,	1,	"01-Apr-24",  "01-Dec-99",	6,	350	,40
,"LPPT",	100,	0,	"01-Jan-11",	"01-Apr-24",1,	40	,100
,"LPPT",	100,	0,	"01-Jan-11",	"01-Apr-24",2,	100	,160
,"LPPT",	100,	0,	"01-Jan-11",	"01-Apr-24",3,	160	,235
,"LPPT",	100,	0,	"01-Jan-11",	"01-Apr-24",4,	235	,295
,"LPPT",	100,	0,	"01-Jan-11",	"01-Apr-24",5,	295	,350
,"LPPT",	100,	0,	"01-Jan-11",	"01-Apr-24",6,	350	,40
,"LPPT",	100,	1,	"01-Apr-24",	"01-Dec-99",1,	40	,100
,"LPPT",	100,	1,	"01-Apr-24",	"01-Dec-99",2,	100	,150
,"LPPT",	100,	1,	"01-Apr-24",	"01-Dec-99",3,	150	,240
,"LPPT",	100,	1,	"01-Apr-24",	"01-Dec-99",4,	240	,300
,"LPPT",	100,	1,	"01-Apr-24",	"01-Dec-99",5,	300	,350
,"LPPT",	100,	1,	"01-Apr-24",	"01-Dec-99",6,	350	,40
)

patch_sectors_overrun_north <- function(.sects_to_fm_bearings){
  this_sects <- .sects_to_fm_bearings
  sects_ok   <- this_sects |> filter(FROM_BEARING <= TO_BEARING)
  sects_nok  <- this_sects |> filter(FROM_BEARING > TO_BEARING)
  if(nrow(sects_nok)){
    sects_nok <- sects_nok|> slice(rep(1:n(), each = 2)) |> 
       mutate( TO_BEARING   = ifelse(row_number() == 1, 360, TO_BEARING)
              ,FROM_BEARING = ifelse(row_number() == 2, 0  , FROM_BEARING)
             )
  }
  return(bind_rows(sects_ok, sects_nok))  # will not bind anything if no overrun
}

lppt_asma40_sects <- lppt_sects |> 
  filter(VERSION == max(VERSION), ASMA_RADIUS == 40) |> 
  patch_sectors_overrun_north()
lppt_asma100_sects <- lppt_sects |> 
  filter(VERSION == max(VERSION), ASMA_RADIUS == 100) |> 
  patch_sectors_overrun_north()

# append sectors to asmas
asmas <- asma_dsma |> filter(PHASE == "ARR")

by40 <- join_by(C40_BEARING >= FROM_BEARING, C40_BEARING < TO_BEARING)
asmas <- asmas |> 
  left_join(lppt_asma40_sects |> 
              select(SECT_40 = ASMA_SECTOR, FROM_BEARING, TO_BEARING)
            , by = by40
            ) |> 
  select(-c(FROM_BEARING, TO_BEARING))

append_asma_100_sectors <- function(.asmas, .sects = lppt_asma100_sects){
  by100 <- join_by(C100_BEARING >= FROM_BEARING, C100_BEARING < TO_BEARING)
  this_asmas <- .asmas |> 
    left_join(.sects |> 
              select(SECT_100 = ASMA_SECTOR, FROM_BEARING, TO_BEARING)
            , by = by100
            ) |> 
  select(-c(FROM_BEARING, TO_BEARING))
  return(this_asmas)
}
```

```{r}
refs_40 <- asmas |> 
  filter(year(MVT_TIME) == 2024) |> 
  group_by(PHASE, CLASS, RWY, SECT_40) |> 
  reframe(N_VALID = sum(!is.na(C40_TRAVEL))
          , REF_20_40 = quantile(C40_TRAVEL, p = 0.2, na.rm = TRUE)
          )

refs_100 <- asmas |> 
  filter(year(MVT_TIME) == 2024) |> 
  group_by(PHASE, CLASS, RWY, SECT_100) |> 
  reframe(N_VALID = sum(!is.na(C100_TRAVEL))
          , REF_20_100 = quantile(C100_TRAVEL, p = 0.2, na.rm = TRUE)
          )
# correct for runway change
refs_100_2 <- refs_100 |> 
  mutate(
    RWY = case_when(
      RWY == "02" ~ "03"
    , RWY == "20" ~ "21"
    , .default = NA
    )
  )
refs_100 <- refs_100 |> bind_rows(refs_100_2)
```

```{r}
asmas <- asmas |> append_asma_100_sectors() 
asmas <- asmas |> left_join(refs_100 |> select(-N_VALID), by = join_by(PHASE, CLASS, RWY, SECT_100))
asmas <- asmas |> mutate(ADD_TIME_100 = C100_TRAVEL - REF_20_100)
```

```{r}
asma_lppt <- asmas |> 
  group_by(ICAO = ADES, DATE, PHASE, CLASS, RWY, SECT_100) |> 
  reframe(ARRS = n()
          , ARRS100 = sum(!is.na(ADD_TIME_100))
          , TOT_REF100 = sum(REF_20_100[!is.na(ADD_TIME_100)])
          , TOT_A100 = sum(C100_TRAVEL[!is.na(ADD_TIME_100)])
          , ADD100 = sum(ADD_TIME_100[!is.na(ADD_TIME_100)]) 
          )
```

```{r}
asma_daily <- asma_lppt |> 
  group_by(ICAO, DATE, PHASE) |> 
  reframe( ARRS100 = sum(ARRS100)
          , ADD100 = sum(ADD100)
          , TOT_A100   = sum(TOT_A100)) |> 
  mutate(AVG_ADD_TIME = ADD100 / ARRS100)
```

asma_daily |> write_csv("./data/EUR-ASMA-LPPT.csv")

ICAO,DOF,ARRS,ARRS40,ARRS100,TOT_A40,TOT_REF40,ADD40,TOT_A100,TOT_REF100,ADD100,AVG_ADD40,AVG_ADD100